---
title: "Features_overview"
author:
- Casper de Visser^[Radboud university medical center, Casper.deVisser@radboudumc.nl]
output:
  html_document:
    toc: true
    df_print: paged
#bibliography: references.bib
params:
  config_file: NULL
  references: NULL
  mae_hdf5_dir_path: NULL
---


```{r}
if (paste0(R.Version()$major, ".", R.Version()$minor) != "4.2.1") {
  warning("This R notebook was implemented using R version 4.2.1")
}
print(params)
config <- yaml::read_yaml(params$config)
outdir <- config$output_directory$FEATURES
#dir.create(outdir)
```

## Read input data 

Input data is a MultiAssayExperiment (MAE) object. 
```{r}
library(MultiAssayExperiment)
library(HDF5Array)

mae <- loadHDF5MultiAssayExperiment(dir = params$mae_hdf5_dir_path, 
                                     prefix = config$mae_hdf5_prefix)

```


## Get omics names list
```{r}
assay_names <- config$experiment_names$CV

# Mapping of original to readable names
assay_name_map <- config$readable_names

# Translate assay names using the map
shorter_assay_names <- sapply(assay_names, function(x) {
  if (!is.null(assay_name_map[[x]])) {
    assay_name_map[[x]]
  } else {
    x  # fallback to original name
  }
})


# Create a data frame for reference
assay_names_df <- data.frame(
  original = assay_names,
  readable = shorter_assay_names,
  stringsAsFactors = FALSE
)

print(assay_names_df)
```



## Functions used to retrieve data from MAE
```{r}

# Function to read omics data from MAE into a scaled dataframe

read_omics_from_mae <- function(mae_object, name_assay){
  se <- mae_object[[name_assay]]
  df_wide <- data.frame(t(assay(se)))
  #df_wide <- df_wide[ , colSums(is.na(df_wide))==0]
  df_wide
  
}

# Function to combine metabolomics data
combine_mtblmcs_data <- function(list_of_assays) {
  df <- cbind(data.frame(list_of_assays$`Very long chain fatty acids | unscaled`),
                      data.frame(list_of_assays$`Acylcarnitines | unscaled`),
                      data.frame(list_of_assays$`Amino acids | unscaled`))
  return(df)
}

combine_mtblmcs_data_sex <- function(list_of_assays, item) {
  df <- cbind(data.frame(list_of_assays$`Very long chain fatty acids | unscaled`[[item]]),
                      data.frame(list_of_assays$`Acylcarnitines | unscaled`[[item]]),
                      data.frame(list_of_assays$`Amino acids | unscaled`[[item]]))
  return(df)
}
```


## Make list of assays
```{r}
assay_list <- lapply(assay_names, function(assay_name) {
  df_wide <- read_omics_from_mae(mae, assay_name)
  }) 
names(assay_list) <- assay_names

# Combine targeted metabolomics, append to assay_list
mtblmcs_combined <- combine_mtblmcs_data(assay_list)

assay_list$`Targeted metabolomics` <- mtblmcs_combined

# Add shorter names 
names(assay_list) <- c(assay_names_df$readable, 'Metabolomics')
```


## Get Omics list, sex stratified
```{r}

# Make list with omics data, for male/female separate

assay_list_sex_stratified <- lapply(assay_names, function(assay_name) {
    # Make separate MAE for Male/Female  
    mae_male <- mae[, colData(mae)$Sex %in% c('MALE')]
    mae_female <- mae[, colData(mae)$Sex %in% c('FEMALE')]
    
    # Read omics data into dataframes
    df_male <- read_omics_from_mae(mae_male, assay_name)
    df_female  <- read_omics_from_mae(mae_female, assay_name)
    
    list(df_male, df_female)
  }) 

# Combine targeted metabolomics, append to assay_list
mtblmcs_combined_male <- combine_mtblmcs_data_sex(assay_list_sex_stratified, 1)
mtblmcs_combined_female <- combine_mtblmcs_data_sex(assay_list_sex_stratified, 2)


assay_list_sex_stratified$`Targeted metabolomics` <- list(mtblmcs_combined_male, mtblmcs_combined_female)

# Add shorter names 
names(assay_list_sex_stratified) <- c(assay_names_df$readable, 'Metabolomics')
```


## Functions to calculate reference values (CV, SD, MEAN, MEDIAN, MIN and MAX)

```{r, fig.width = 20}
library(ggplot2)
library(ggrepel)
library(ggokabeito) # Okabe-Ito


# Function to create dataframe of Coefficients of Variance (CV) per feature
make_CV_df <- function(df_input) {
  
  # Calculate Coefficients of Variance
  df_input <- df_input + (abs(min(df_input, na.rm = T)) + 1)
  CV_list <- sapply(df_input, function(x) sd(x, na.rm=T) / mean(x, na.rm=T) * 100)
  #CV_list <- sapply(df_input, function(x) sd(x))
  
  df_out <- data.frame(CV_list)
  colnames(df_out) <- c('CV')
  
  return(df_out)
}


make_ref_values_df <- function(df_input) {
  
  # Calculate Coefficients of Variance
  df_input_cv <- df_input + (abs(min(df_input, na.rm =T)) + 1)
  CV_list <- sapply(df_input_cv, function(x) sd(x, na.rm=T) / mean(x, na.rm=T) * 100)
  
  # Calculate more reference values
  SD_list <- sapply(df_input, function(x) abs(sd(x, na.rm=T)))
  mean_list <- sapply(df_input, function(x) mean(x, na.rm=T))
  median_list <- sapply(df_input, function(x) median(x, na.rm=T))
  min_list <- sapply(df_input, function(x) min(x, na.rm=T))
  max_list <- sapply(df_input, function(x) max(x, na.rm=T))

  
  df_out <- data.frame(CV_list, SD_list,
                       mean_list, median_list,
                       min_list, max_list)
  colnames(df_out) <- c('CV','SD', 'MEAN', 'MEDIAN', 'MIN', 'MAX')
  
  return(df_out)
}


```


## Make density plots of CVs for each -omics, sex-stratified
```{r, fig.width=20}
# Function to make density plot for sex-stratified data

save_reference_values2csv <- function(item) {
  
  # Get Data frames, omics name
  input_df_male <- assay_list_sex_stratified[[item]][[1]]
  input_df_female <- assay_list_sex_stratified[[item]][[2]]
  name <- names(assay_list_sex_stratified)[[item]]
  print(name)
  
  # Male
  Ref_df_male <- make_ref_values_df(input_df_male)
  
  # Female
  Ref_df_female <- make_ref_values_df(input_df_female)

  # Dataframe to save as CSV
  df_combined <- merge(Ref_df_male, Ref_df_female, by = 'row.names', all = T)
  
  colnames(df_combined) <- c('Feature',
                             'CV_male', 'SD_male',
                             'MEAN_male', 'MEDIAN_male',
                             'MIN_male', 'MAX_male', 
                             'CV_Female', 'SD_Female',
                             'MEAN_Female', 'MEDIAN_Female',
                             'MIN_Female', 'MAX_Female')

   # Save to CSV
  outdir <- config$output_directory$CV
  filename <- paste0(name, "_CVs.csv")
  out_path <- file.path(outdir, filename)
  dir.create(out_path)
  write.csv(df_combined, filename, row.names = F)
}


make_CV_plot_sex_stratified <- function(item) {
  
  # Get Data frames, omics name
  input_df_male <- assay_list_sex_stratified[[item]][[1]]
  input_df_female <- assay_list_sex_stratified[[item]][[2]]
  name <- names(assay_list_sex_stratified)[[item]]
  
  # Male
  CVs_df_male <- make_CV_df(input_df_male)
  CVs_df_male$Sex <- rep(paste0('Male'), times=nrow(CVs_df_male))
  
  # Female
  CVs_df_female <- make_CV_df(input_df_female)
  CVs_df_female$Sex <- rep(paste0('Female'), times=nrow(CVs_df_female))

  # Combine CV dataframes into one for plot
  Df_combined <- rbind(CVs_df_male, CVs_df_female)

  p <- ggplot(Df_combined, aes(x=CV, fill=Sex)) +
    geom_density(color = 'darkblue', alpha=0.7)+
    theme_bw() +
    scale_fill_okabe_ito(order = c(7,5)) +
    ggtitle(paste0(name, ", ", nrow(CVs_df_male), " features")) + 
    theme_grey(base_size=22) +
    theme(plot.title = element_text(size = 25)) +
    xlim(c(0,40))
  return(p)
}
```


## Save reference values to .csv files
```{r}
lapply(seq(1:nrow(assay_names_df)), function(x) save_reference_values2csv(x))
```

## Plot Coefficients of Variances per -omics type
```{r, fig.width=20}
lapply(seq(1:nrow(assay_names_df)), function(x) make_CV_plot_sex_stratified(x)) ##Plot all
plot_list <- lapply(1:nrow(assay_names_df), function(x) make_CV_plot_sex_stratified(x)) ## Without aCGH
```

### Histogram for Very Long Chain Fatty Acids

```{r, fig.width=20}
input_df_male <- assay_list_sex_stratified[["VLCFA"]][[1]]
input_df_female <- assay_list_sex_stratified[["VLCFA"]][[2]]
  
# Male
CVs_df_male <- make_CV_df(input_df_male)
CVs_df_male$Sex <- rep(paste0('Male'), times=nrow(CVs_df_male))
  
# Female
CVs_df_female <- make_CV_df(input_df_female)
CVs_df_female$Sex <- rep(paste0('Female'), times=nrow(CVs_df_female))

# Combine dataframes into onec
Df_combined <- rbind(CVs_df_male, CVs_df_female)
colnames(Df_combined) <- c('CV', 'Sex')

p <- ggplot(Df_combined, aes(x=CV, fill=Sex)) +
    geom_histogram(color = 'darkblue', bins = 60) +
    #theme_bw() +
    
    scale_fill_okabe_ito(order = c(7,5)) +
    ggtitle('Very Long Chain Fatty Acids, 5 features') + 
    theme_grey(base_size=22) +
    theme(plot.title = element_text(size = 25)) +
    xlim(c(0,40))
p
```

## Grid arrange all plots together
```{r}
library(gridExtra)

plot_list[[4]] <- p
n <- length(plot_list)
nrow <- ceiling(n /2)
pdf("arranged_CVplot.pdf", width = 26, height = 3 * nrow)
do.call(grid.arrange, c(plot_list, ncol = 2, padding=unit(20, "cm")))
dev.off()
```

