---
title: 'Robust linear models -- preprocessed single omics data'
author:
  - Casper de Visser^[Radboud university medical center, Anna.Niehues@radboudumc.nl]
output:
  html_document:
    toc: true
    df_print: paged
params:
  config_file: NULL
  references: NULL
  mae_hdf5_dir_path: NULL
---

This R notebook performs robust linear modeling per omics feature (molecualr entity). Robust models [@Huber1981;@Hampel1986;@Marazzi1993;@Venables2022] are implemented using the *R* package `MASS` [@Venables2022].

```{r}
if (paste0(R.Version()$major, ".", R.Version()$minor) != "4.2.1") {
  warning("This R notebook was implemented using R version 4.2.1")
}
print(params)
config <- yaml::read_yaml(params$config)
outdir <- config$output_directory$RLM
dir.create(outdir)
```

## Read data and metadata: MultiAssayExperiment

The `MultiAssayExperiment` contains preprocessed data and has been created with the R notebook https://gitlab.cmbi.umcn.nl/eatris-plus/mtblm_data_preparation/-/blob/main/R_data_preprocessing/preprocessing_targeted_metabolomics_data/Rmd. Specific assays/experiments used in this analysis are defined below.

```{r read_multiassayexperiment}
library(MultiAssayExperiment)

mae <- loadHDF5MultiAssayExperiment(
  dir = config$mae_hdf5_dir_path, 
  prefix = config$mae_hdf5_prefix)
mae
```


Select assays for analysis
```{r}
## Get omics names list
assay_names <- config$experiment_names$RLM

# Mapping of original to readable names
assay_name_map <- config$readable_names

# Translate assay names using the map
shorter_assay_names <- sapply(assay_names, function(x) {
  if (!is.null(assay_name_map[[x]])) {
    assay_name_map[[x]]
  } else {
    x  # fallback to original name
  }
})


# Create a data frame for reference
assay_names_df <- data.frame(
  original = assay_names,
  readable = shorter_assay_names,
  stringsAsFactors = FALSE
)
```


```{r}
library(MASS)
library(sfsmisc)
```


## RMSE of examples for manuscript
```{r}
run_rlm_df <- function(feature, df, outcome, covariates = NULL) {
    # run rlm for one feature and one outcome
    df$y <- df[[feature]]
    # define formula
    covar_term <- ""
    if (!is.null(covariates)) {
      covar_term <- paste0(" + ", paste0(covariates, collapse = " + "))}
    str_formula <- paste0("y ~ ", outcome, covar_term)
    my_formula <- as.formula(str_formula)
    # fit robust regression model
    rlm_fit <- try(rlm(formula = my_formula, data = df, maxit = 50))
    if (length(rlm_fit) < 2) {
      message("MASS::rlm error: ", feature, ", ", str_formula)
      rlm_results = data.frame(
        feature = feature,
        outcome = outcome,
        formula = str_formula,
        covariates = paste0(covariates, collapse = "+"),
        coefficient = NA, 
        std.error = NA,
        t.value = NA,
        p.value = NA)
    } else {
      # robust F-test - Wald test for multiple coefficients of rlm object
      f_test <- f.robftest(rlm_fit, var = 2) # 1=intercept, 3:=covariates
      # results
      rlm_results = data.frame(
        feature = feature,
        outcome = outcome,
        formula = str_formula,
        covariates = paste0(covariates, collapse = "+"),
        coefficient = summary(rlm_fit)$coefficients[2, "Value"], 
        std.error = summary(rlm_fit)$coefficients[2, "Std. Error"],
        t.value = summary(rlm_fit)$coefficients[2, "t value"],
        p.value = f_test$p.value)
    }
}
```


## Tigclycarnitine
```{r}
se <- mae[["Acylcarnitines | batch-adjusted | with missing values"]]
df = as.data.frame(t(assays(se)[[1]]))
df <- cbind(df, colData(mae)[rownames(df), ]) 
outcome = 'ABO.Blood.Group'
covariates =  NULL
feature <- "Tiglylcarnitine"

df$y <- df[,feature]
str_formula <- paste0("y ~ ", outcome, NULL)
my_formula <- as.formula(str_formula)
model <- rlm_fit <- try(rlm(formula = my_formula, data = df, maxit = 50))
residuals <- resid(model)
sqrt(mean(residuals^2))

```

```{r}
se <- mae[["Amino acids | batch-adjusted | with missing values"]]
df = as.data.frame(t(assays(se)[[1]]))
df <- cbind(df, colData(mae)[rownames(df), ]) 
outcome = 'Sex'
covariates =  NULL
feature <- "O-Phosphoethanolamine"

df$y <- df[,feature]
str_formula <- paste0("y ~ ", outcome, NULL)
my_formula <- as.formula(str_formula)
model <- rlm_fit <- try(rlm(formula = my_formula, data = df, maxit = 50))
residuals <- resid(model)
sqrt(mean(residuals^2))
```
# Power analysis (aging features as examples)
```{r}
library(pwr)
library(ggplot2)
```


Plot function for smaller cohort sizes
```{r}
plot_power_curve <- function(
    mae,
    assay_name,
    outcome,
    feature,
    covariates = NULL,
    alpha = 0.05,
    sample_seq = NULL,
    min_n = 20,
    by = 5,
    return_data = FALSE
) {

    # extract assay
    se <- mae[[assay_name]]

    # build dataframe
    df <- as.data.frame(t(assays(se)[[1]]))
    df <- cbind(df, colData(mae)[rownames(df), ])

    # response variable
    df$y <- df[, feature]

    # build formula
    covar_term <- ""

    if (!is.null(covariates)) {
        covar_term <- paste0(
            " + ",
            paste0(covariates, collapse = " + ")
        )
    }

    str_formula <- paste0(
        "y ~ ",
        outcome,
        covar_term
    )

    my_formula <- as.formula(str_formula)

    # fit linear model
    lm_fit <- lm(my_formula, data = df)

    # effect size
    R2 <- summary(lm_fit)$r.squared
    f2 <- R2 / (1 - R2)

    # number of predictors
    u <- 1

    if (!is.null(covariates)) {
        u <- u + length(covariates)
    }

    # sample sizes
    if (is.null(sample_seq)) {
        sample_sizes <- c(
            seq(min_n, nrow(df), by = by),
            nrow(df)
        )

        sample_sizes <- unique(sample_sizes)
    } else {
        sample_sizes <- sample_seq
    }

    # compute power
    powers <- sapply(sample_sizes, function(n) {

        v <- n - u - 1

        pwr.f2.test(
            u = u,
            v = v,
            f2 = f2,
            sig.level = alpha
        )$power
    })

    # dataframe for plotting
    power_df <- data.frame(
        sample_size = sample_sizes,
        power = powers
    )

    # plot
    p <- ggplot(power_df,
                aes(x = sample_size,
                    y = power)) +

        geom_line(linewidth = 1.2) +

        geom_hline(
            yintercept = 0.80,
            linetype = "dashed"
        ) +

        coord_cartesian(ylim = c(0,1)) +

        labs(
            x = "Sample size",
            y = "Statistical power",
            title = paste0(outcome, " ~ ", feature),
            subtitle = assay_name
        ) +

        theme_bw(base_size = 13) +

        theme(
            plot.title = element_text(face = "bold"),
            axis.title = element_text(face = "bold")
        )

    # return
    if (return_data) {

        return(list(
            plot = p,
            power_df = power_df,
            lm_fit = lm_fit,
            R2 = R2,
            f2 = f2
        ))

    } else {

        return(p)
    }
}
```


## Examples for biomarkers of biological age
```{r}
p <- plot_power_curve(
  mae = mae,
  assay_name = "miRNA-seq, mature | batch-adjusted | corrected",
  outcome = 'Age_pred_Levine', 
  feature = "hsa.miR.6876.3p_MIMAT0027653_Homo_sapiens_miR.6876.3p",
)
file.out <- file.path(outdir, 'miRNAseq_AgePredLevine_miR-6876-3p.pdf')
ggsave(filename = file.out, p, width = 8, height = 5.5)
```

```{r}
p <- plot_power_curve(
  mae = mae,
  assay_name = "Lipidomics, positive | transformed",
  outcome = 'Age_acc_Levine', 
  feature = "M637T785",
)
file.out <- file.path(outdir, 'Lipid_AgeAccLevine_M637T785.pdf')
ggsave(filename = file.out, p, width = 8, height = 5.5)
```

```{r}
p <- plot_power_curve(
  mae = mae,
  assay_name = "Proteomics | imputed missing values",
  outcome = 'Age_acc_BLUP', 
  feature = "A0A5C2GF79",
)
file.out <- file.path(outdir, 'Prot_AgeAccBLUP_A0A5C2GF79.pdf')
ggsave(filename = file.out, p, width = 8, height = 5.5)
```


## Examples of miRNAs from Huan et al. (2017)
```{r}
p <- plot_power_curve(
  mae = mae,
  assay_name = "miRNA-seq, mature | batch-adjusted | corrected",
  outcome = 'Age', 
  feature = "hsa.miR.99b.5p_MIMAT0000689_Homo_sapiens_miR.99b.5p",
)
file.out <- file.path(outdir, 'miRNAseq_ChronAge_miR-99b-5p.pdf')
ggsave(filename = file.out, p, width = 8, height = 5.5)
```

```{r}
p <- plot_power_curve(
  mae = mae,
  assay_name = "miRNA-seq, mature | batch-adjusted | corrected",
  outcome = 'Age', 
  feature = "hsa.miR.130b.5p_MIMAT0004680_Homo_sapiens_miR.130b.5p",
)
file.out <- file.path(outdir, 'miRNAseq_ChronAge_miR-130b-5p.pdf')
ggsave(filename = file.out, p, width = 8, height = 5.5)
```

```{r}
p <- plot_power_curve(
  mae = mae,
  assay_name = "miRNA-seq, mature | batch-adjusted | corrected",
  outcome = 'Age', 
  feature = "hsa.miR.302c.3p_MIMAT0000717_Homo_sapiens_miR.302c.3p",
)
file.out <- file.path(outdir, 'miRNAseq_ChronAge_miR-302c-3p.pdf')
ggsave(filename = file.out, p, width = 8, height = 5.5)
```


## Session information

```{r}
sessionInfo()
```

## References

