---
title: 'Robust linear models -- preprocessed single omics data'
author:
  - Anna Niehues^[Radboud university medical center, Anna.Niehues@radboudumc.nl]
output:
  html_document:
    toc: true
    df_print: paged
params:
  config_file: NULL
  references: NULL
  mae_hdf5_dir_path: NULL
---

This R notebook performs robust linear modeling per omics feature (molecualr entity). Robust models [@Huber1981;@Hampel1986;@Marazzi1993;@Venables2022] are implemented using the *R* package `MASS` [@Venables2022].

```{r}
if (paste0(R.Version()$major, ".", R.Version()$minor) != "4.2.1") {
  warning("This R notebook was implemented using R version 4.2.1")
}
print(params)
config <- yaml::read_yaml(params$config)
outdir <- config$output_directory$RLM
dir.create(outdir)
```

## Read data and metadata: MultiAssayExperiment

The `MultiAssayExperiment` contains preprocessed data and has been created with the R notebook https://gitlab.cmbi.umcn.nl/eatris-plus/mtblm_data_preparation/-/blob/main/R_data_preprocessing/preprocessing_targeted_metabolomics_data/Rmd. Specific assays/experiments used in this analysis are defined below.

```{r read_multiassayexperiment}
library(MultiAssayExperiment)

mae <- loadHDF5MultiAssayExperiment(
  dir = config$mae_hdf5_dir_path, 
  prefix = config$mae_hdf5_prefix)
mae
```

The MultiAssayExperiment contains individual targeted metabolomics assays

```{r assays_overview}
experiments(mae)
```

Select assays for analysis
```{r}
## Get omics names list
assay_names <- config$experiment_names$RLM

# Mapping of original to readable names
assay_name_map <- config$readable_names

# Translate assay names using the map
shorter_assay_names <- sapply(assay_names, function(x) {
  if (!is.null(assay_name_map[[x]])) {
    assay_name_map[[x]]
  } else {
    x  # fallback to original name
  }
})


# Create a data frame for reference
assay_names_df <- data.frame(
  original = assay_names,
  readable = shorter_assay_names,
  stringsAsFactors = FALSE
)
```


Technical variables are available per assay

```{r assay_metadata_overview}
lapply(experiments(mae), function(x) {names(colData(x))})
```


Phenotype information shared across all assays can be accessed via `colData()`

```{r sample_information_overview}
names(colData(mae))
```

Add variable that captures both sex and phenotypic feature

```{r}
# variable describing both Sex and PhenotypicFeature
colData(mae)$Sex.PhenotypicFeature <- gsub(
  ".NA", "", paste(colData(mae)$Sex, colData(mae)$PhenotypicFeature, sep = "."))
```


## Feature-wise robust linear models

Feature-wise models using robust regression with `MASS::rlm` [@Huber1981;@Hampel1986;@Marazzi1993;@Venables2022]. For types of models are applied: 

1. Using all samples without covariate
    * Omics_feature ~ biological_variable
2. Using all samples, including biological confounder (sex) as covariate
    * Omics_feature ~ biological_variable + sex
3. Using samples from male individuals
    * Omics_feature ~ biological_variable
4. Using samples from female individuals
    * Omics_feature ~ biological_variable

The models are applied per biological variable, i.e., phenotype information (age, BMI, etc.) or clinical hematology variable (hemoglobin, lymphocytes, etc.).

Errors can occur when data is very sparse (e.g., miRNA-seq data).

```{r}
library(MASS)
library(sfsmisc)
```

```{r robust_linear_modeling, include = FALSE}
#' Apply robust linear modeling to one assay and multiple outcome variables
#'
#' @param se A SummarizedExperiment from experiments(MultiAssayExperiment).
#' @param outcomes A character vector with column names present in colData(MultiAssayExperiment).
#' @param covariates A character vector with column names present in colData(MultiAssayExperiment).
#' @returns A data frame.
#' @examples
#' run_rlm_se(se, outcomes = c("Age", "BMI"), covariates = c("Sex"))
run_rlm_se <- function(se, outcomes, covariates = NULL) {
  print(se)
  # run rlm on SummarizedExperiment
  require(MASS)
  require(sfsmisc)
  # create data frame with omics feature values
  df <- as.data.frame(t(assays(se)[[1]]))
  # remove non-variable features
  print(dim(df))
  df <- df[, apply(df, 2, function(column) {
    var(column, na.rm = T) != 0 })]
  # remove extremely sparse features
  xpercent <- nrow(df)*0.5
  df <- df[, apply(df, 2, function(column) {
    sum(column == 0, na.rm = T) < xpercent })]
  print(dim(df))
  features <- colnames(df)
  # scale and center feature columns
  #df <- as.data.frame(scale(df, center = T, scale = T))
  print(head(df))
  # add biological variables to data frame
  df <- cbind(df, colData(mae)[rownames(df), ]) 
  # --> TODO: add mae as parameter and use experiment name to select se
  # add assay-specific technical variables to data frame
  df <- cbind(df, colData(se))
  
  #' Apply robust linear modeling to one feature and one outcome variable
  #'
  #' @param feature A molecule identifier present in rowData(SummarizedExperiment).
  #' @param df A data frame containing assay data and biological variables.
  #' @param outcome A character string indicating biological variable column name.
  #' @param covariates A character string indicating covariate column name.
  #' @returns A data frame.
  #' @examples
  #' run_rlm_df("hmdb:HMDB0000062", df, outcome = "Age", covariates = c("Sex"))
  run_rlm_df <- function(feature, df, outcome, covariates = NULL) {
    # run rlm for one feature and one outcome
    df$y <- df[[feature]]
    # define formula
    covar_term <- ""
    if (!is.null(covariates)) {
      covar_term <- paste0(" + ", paste0(covariates, collapse = " + "))}
    str_formula <- paste0("y ~ ", outcome, covar_term)
    my_formula <- as.formula(str_formula)
    # fit robust regression model
    rlm_fit <- try(rlm(formula = my_formula, data = df, maxit = 50))
    if (length(rlm_fit) < 2) {
      message("MASS::rlm error: ", feature, ", ", str_formula)
      rlm_results = data.frame(
        feature = feature,
        outcome = outcome,
        formula = str_formula,
        covariates = paste0(covariates, collapse = "+"),
        coefficient = NA, 
        std.error = NA,
        t.value = NA,
        p.value = NA)
    } else {
      # robust F-test - Wald test for multiple coefficients of rlm object
      f_test <- f.robftest(rlm_fit, var = 2) # 1=intercept, 3:=covariates
      # results
      rlm_results = data.frame(
        feature = feature,
        outcome = outcome,
        formula = str_formula,
        covariates = paste0(covariates, collapse = "+"),
        coefficient = summary(rlm_fit)$coefficients[2, "Value"], 
        std.error = summary(rlm_fit)$coefficients[2, "Std. Error"],
        t.value = summary(rlm_fit)$coefficients[2, "t value"],
        p.value = f_test$p.value)
    }
  }
  
  names(outcomes) <- outcomes
  
  rlm_results <- do.call(rbind, lapply(outcomes, function(outcome) {
    tmp_df <- df
    if (length(outcome) > 50) {
          outcome = "All"
        }
    else {
        outcome = outcome
    }
    # remove missing values categories
    if (outcome == "Smoking.Status") {
      tmp_df <- tmp_df[tmp_df$Smoking.Status != "Unknown If Ever Smoked", ]
      tmp_df$Smoking.Status <- droplevels(tmp_df$Smoking.Status)
    }
   
    rlm_result_ <- do.call(rbind, lapply(
      features,
      run_rlm_df,
      df = tmp_df, outcome = outcome, covariates = covariates))
    # adjust p-values
    rlm_result_$p.adj <- p.adjust(rlm_result_$p.value, method = "BH")
    rlm_result_$rank <- rank(rlm_result_$p.value)
    rlm_result_
  }))
  rlm_results
}


# outcome variables
outcomes <- c(
  "Sex", "Age", "Age.group", 
  "Height", "Weight", "BMI", "BMI.group", 
  "Smoking.Status", "ABO.Blood.Group", "Rh.Blood.Group", 
  "Leukocytes", "Erythrocytes", "Hemoglobin", "Hematocrit", "Platelets", "MCV", 
  "MCH", "MCHC", "Lymphocytes", "MXD", "Neutrophils",
  "Age_pred_BLUP", "Age_acc_BLUP",   "Age_pred_Levine", "Age_acc_Levine")

# TODO: Look at combined outcome if possible
# Presumably, there are just no significant associations when combing these
# outcomes <- c(
#   "Sex", "Age", "Age.group", 
#   "Height", "Weight", "BMI", "BMI.group", 
#   "Smoking.Status", "ABO.Blood.Group", "Rh.Blood.Group", 
#   "Leukocytes", "Erythrocytes", "Hemoglobin", "Hematocrit", "Platelets", "MCV", 
#   "MCH", "MCHC", "Lymphocytes", "MXD", "Neutrophils", 
#   "Sex + Age + Age.group + Height + Weight + BMI + BMI.group + Smoking.Status + ABO.Blood.Group + Rh.Blood.Group +
#   Leukocytes + Erythrocytes + Hemoglobin + Hematocrit + Platelets + MCV + 
#   MCH + MCHC + Lymphocytes + MXD + Neutrophils")

write_last_column_smaller <- function(df_input) {
  df_input$outcome[df_input$outcome == "Sex + Age + Age.group + Height + Weight + BMI + BMI.group + Smoking.Status + ABO.Blood.Group + Rh.Blood.Group +
  Leukocytes + Erythrocytes + Hemoglobin + Hematocrit + Platelets + MCV + 
  MCH + MCHC + Lymphocytes + MXD + Neutrophils"] <- "All"
  return(df_input)
}


# fit robust linear models using different formulas
mae_rlm_woCovariate <- lapply( # without any covariates
  experiments(mae)[assay_names], run_rlm_se,
  outcomes = outcomes)

mae_rlm_wCovarSex <- lapply( # include sex as covariate
  experiments(mae)[assay_names], run_rlm_se,
  outcomes = outcomes[outcomes != "Sex"],
  covariates = c("Sex"))

mae_rlm_woCovariate_m <- lapply( # male only
  experiments(mae[, colData(mae)$Sex == "MALE", ])[assay_names], 
  run_rlm_se, 
  outcomes = outcomes[-which(outcomes %in% c("Sex"))])

mae_rlm_woCovariate_f <- lapply( # female only
  experiments(mae[, colData(mae)$Sex == "FEMALE", ])[assay_names], 
  run_rlm_se, 
  outcomes = outcomes[which(outcomes != "Sex")])


# merge all rlm results into one list
mae_rlm_all <- list(
  `w/o covariates` = mae_rlm_woCovariate, 
  `with sex as covariates` = mae_rlm_wCovarSex,
  `male only, w/o covariates` = mae_rlm_woCovariate_m, 
  `female only, w/o covariates` = mae_rlm_woCovariate_f)
```

Print top results

```{r}
# top 10 features across all outcomes
lapply(names(mae_rlm_all), function(method_name) {
  lapply(assay_names, function(assay_name) {
    rlm_res <- mae_rlm_all[[method_name]][[assay_name]]
    head(rlm_res[order(rlm_res$p.value, decreasing = FALSE),], n = 3)
  })
})
```

### Visualize results

```{r}
library(ggplot2)
library(ggrepel)
library(ggokabeito) # Okabe-Ito
```

Plot histograms of p-values per assay

```{r pval_histograms}
# p-values histograms
lapply(assay_names, function(assay_name) {
  lapply(names(mae_rlm_all), function(method_name) {
    rlm_res <- na.omit(mae_rlm_all[[method_name]][[assay_name]])
    if (dim(rlm_res)[1] > 0) { # male only, fatty acids, yielded no results in rlm
      ggplot(rlm_res, aes(x = p.value)) +
        geom_histogram(aes(fill = p.adj<=0.05), bins = 30) +
        facet_wrap(vars(outcome)) +
        scale_color_okabe_ito() + scale_fill_okabe_ito() +
        ggtitle(stringr::str_wrap(
          paste0(assay_name, ", robust linear models, ", method_name),
          width = 60))
    }
  })
})
```

Volcano plots per phenotypic variable

```{r volcano_plots, fig.width = 11, fig.height = 4.5}
# volcano plots
mae_rlm_all_merged <- do.call(
  rbind, lapply(names(mae_rlm_all), function(method_name) {
    rlm_res_method <- do.call(rbind, lapply(assay_names, function(assay_name) {
      rlm_res_ <- mae_rlm_all[[method_name]][[assay_name]]
      rlm_res_$assay <- assay_name #strsplit(assay_name, "_")[[1]][[2]]
      
      if (startsWith(assay_name, "Lipidomics")) {
        rlm_res_$feature2 <- data.frame(rowData(experiments(mae)[[assay_name]]))[
          rlm_res_$feature, ][["Name"]]
      } else {
        rlm_res_$feature2 <- rlm_res_$feature
      }
      # label only significant; limit to top x significant
      rlm_res_$label <- ifelse((rlm_res_$p.adj<=0.05) & (rlm_res_$rank < 4), 
                               rlm_res_$feature2, NA)
      method_split <- strsplit(method_name, " ")
      rlm_res_$dataset <- apply(rlm_res_, 1, function(x) {
        if (method_split[[1]][1] == "male") {
          dat = "male"
        } else if (method_split[[1]][1] == "female") {
          dat = "female"
        } else {
          dat = "all"
        }
        dat
      })
      rlm_res_
    })) 
  }))


lapply(outcomes, function(outcome) {
  rlm_res <- mae_rlm_all_merged[mae_rlm_all_merged$outcome == outcome, ]
  rlm_res$method <- stringr::str_wrap(paste(rlm_res$dataset, " | ", rlm_res$formula), width = 20)
  # my_levels <- unique(rlm_res$method)[order(sapply(unique(rlm_res$method), nchar))]
  my_levels <- unique(rlm_res$method)[order(unique(rlm_res$method))]
  rlm_res$method <- factor(rlm_res$method, levels = my_levels)
  rlm_res$assay <- factor(rlm_res$assay, levels = assay_names)
  
  ggp <- ggplot(rlm_res,
         aes(x = coefficient, y = -log10(p.adj))) +
    geom_point(aes(col = p.adj<=0.05, shape = p.adj<=0.05), 
               na.rm = T, size = 1) +
    scale_color_okabe_ito() + scale_fill_okabe_ito() +
    facet_grid(assay~method, scales = "free_y") +
    theme_bw() +
    theme(strip.text.y = element_text(angle = 0),
          text = element_text(size = 13),
          panel.spacing = unit(0.1, "lines")) +
    geom_text_repel(
      aes(label = label), size = 2.5, direction = "both",
      min.segment.length = unit(0.1, "lines"),
      max.time = 3, max.iter = 1000000) +
    xlab("Regression coefficient")
    # ggtitle(stringr::str_wrap(
    #   paste0("Robust linear models, ", outcome), width = 60))
  ggsave(file.path(outdir, paste0("RLMvolcano_", outcome, ".png")),
         ggp, width = 28.04, height = 10.46, units = "cm", dpi = 300)
  
  ggp
})
```

Compare number (or proportion) of significant hits per omics, outcome variable, and model.

```{r, fig.width=11, fig.height=6.5}
require(dplyr)
require(tidyr)

names(mae_rlm_all_merged)
mae_rlm_all_merged$significant <- mae_rlm_all_merged$p.adj <= 0.05
mae_rlm_all_merged$method <- as.factor(paste0(
  mae_rlm_all_merged$dataset, 
  ifelse(mae_rlm_all_merged$covariates != "", 
         paste0(" | covariate: ", mae_rlm_all_merged$covariates), 
         " | no covariate")))


# TODO: This should not be used if not using this combined outcome
#mae_rlm_all_merged <- write_last_column_smaller(mae_rlm_all_merged)

# calculate proportion of significant features per omics, outcome, data set/model
df <- mae_rlm_all_merged %>%
  group_by(assay, outcome, method) %>%
  count(significant, name = "n_signif") %>%
  group_by(assay, outcome, method) %>%
  mutate(freq = n_signif / sum(n_signif)) %>%
  drop_na()
```


## Save Significant associations to CSV (THIS WILL BE THE END OF THIS SCRIPT, or after change the assay names?)
```{r}
signif_hits <- mae_rlm_all_merged[mae_rlm_all_merged$p.adj < 0.05,]
# Save dataframe with significant hits
write.csv(signif_hits, file.path(outdir, "signif_hits.csv"))
``` 


## RMSE of examples for manuscript
```{r}
run_rlm_df <- function(feature, df, outcome, covariates = NULL) {
    # run rlm for one feature and one outcome
    df$y <- df[[feature]]
    # define formula
    covar_term <- ""
    if (!is.null(covariates)) {
      covar_term <- paste0(" + ", paste0(covariates, collapse = " + "))}
    str_formula <- paste0("y ~ ", outcome, covar_term)
    my_formula <- as.formula(str_formula)
    # fit robust regression model
    rlm_fit <- try(rlm(formula = my_formula, data = df, maxit = 50))
    if (length(rlm_fit) < 2) {
      message("MASS::rlm error: ", feature, ", ", str_formula)
      rlm_results = data.frame(
        feature = feature,
        outcome = outcome,
        formula = str_formula,
        covariates = paste0(covariates, collapse = "+"),
        coefficient = NA, 
        std.error = NA,
        t.value = NA,
        p.value = NA)
    } else {
      # robust F-test - Wald test for multiple coefficients of rlm object
      f_test <- f.robftest(rlm_fit, var = 2) # 1=intercept, 3:=covariates
      # results
      rlm_results = data.frame(
        feature = feature,
        outcome = outcome,
        formula = str_formula,
        covariates = paste0(covariates, collapse = "+"),
        coefficient = summary(rlm_fit)$coefficients[2, "Value"], 
        std.error = summary(rlm_fit)$coefficients[2, "Std. Error"],
        t.value = summary(rlm_fit)$coefficients[2, "t value"],
        p.value = f_test$p.value)
    }
}
```


## Tigclycarnitine
```{r}
se <- mae[["Acylcarnitines | batch-adjusted | with missing values"]]
df = as.data.frame(t(assays(se)[[1]]))
df <- cbind(df, colData(mae)[rownames(df), ]) 
outcome = 'ABO.Blood.Group'
covariates =  NULL
feature <- "Tiglylcarnitine"

df$y <- df[,feature]
str_formula <- paste0("y ~ ", outcome, NULL)
my_formula <- as.formula(str_formula)
model <- rlm_fit <- try(rlm(formula = my_formula, data = df, maxit = 50))
residuals <- resid(model)
sqrt(mean(residuals^2))

```

```{r}
se <- mae[["Amino acids | batch-adjusted | with missing values"]]
df = as.data.frame(t(assays(se)[[1]]))
df <- cbind(df, colData(mae)[rownames(df), ]) 
outcome = 'Sex'
covariates =  NULL
feature <- "O-Phosphoethanolamine"

df$y <- df[,feature]
str_formula <- paste0("y ~ ", outcome, NULL)
my_formula <- as.formula(str_formula)
model <- rlm_fit <- try(rlm(formula = my_formula, data = df, maxit = 50))
residuals <- resid(model)
sqrt(mean(residuals^2))
```


## Session information

```{r}
sessionInfo()
```

## References

