---
title: 'Principal component analysis (PCA)'
author:
  - Casper de Visser^[Radboud university medical center, Casper.deVisser@radboudumc.nl]
  - Anna Niehues^[Radboud university medical center, Anna.Niehues@radboudumc.nl]
output:
  html_document:
    toc: true
    df_print: paged
#bibliography: references.bib
params:
  config_file: "config.yml"
---

This R notebook performs Principal Component Analysis (PCA) using `prcomp` [@R2022] on each individual omics data set provided in a `MultiAssayExperiment` [@Ramos2017].
Some omics types have different versions of processed data, e.g., data containing missing values and data with imputed missing values.
As PCA cannot handle missing values, data with imputed values is used for this analysis. 
If both batch effect adjusted and unadjusted data are available, PCA is performed on both data sets.

## Read data and metadata: MultiAssayExperiment

The `MultiAssayExperiment` [@Ramos2017] contains preprocessed data and has been created with the *R* notebook `examples\create_MultiAssayExperiment.Rmd`.

```{r read_multiassayexperiment}
if (paste0(R.Version()$major, ".", R.Version()$minor) != "4.2.1") {
  warning("This R notebook was implemented using R version 4.2.1")
}
print(params)
config <- yaml::read_yaml(params$config)
outdir <- config$output_directory$PCA
dir.create(outdir)
```



```{r}
require(SummarizedExperiment)
library(MultiAssayExperiment)


mae <- loadHDF5MultiAssayExperiment(
  dir = config$mae_hdf5_dir_path, 
  prefix = config$mae_hdf5_prefix)
mae
```

The MultiAssayExperiment contains individual omics assays

```{r}
experiments(mae)
```

```{r}
# extract names of experiments
experiment_names <- config$experiment_names$PCA
names(experiment_names) <- experiment_names

# # exclude experiments that contain missing data
# experiment_names <- experiment_names[
#   !sapply(strsplit(experiment_names, " | ", fixed = T), function(n) {
#     "with missing values" %in% n })]

experiment_names

# make sure that there are no missing values
stopifnot(sapply(experiment_names, function(n) {
  !any(is.na(assays(mae[[n]])[[1]]))
}))
```


Technical variables are available per assay

```{r}
tech_vars_factor <- lapply(experiments(mae), function(x) {
  num_levels <- apply(colData(x), 2, function(x) {length(levels(as.factor(x)))})
  t <- names(colData(x))[num_levels <= 9 & num_levels > 1]
  t <- t[t!="subject.id"]
  t <- t[!t %in% names(colData(mae))]
})
tech_vars_factor

tech_vars_continuous <- lapply(experiments(mae), function(x) {
  num_levels <- apply(colData(x), 2, function(x) {length(levels(as.factor(x)))})
  t <- names(colData(x))[num_levels > 9]
  t <- t[t!="subject.id"]
  t <- t[t!="measurement.id"]
  t <- t[t!="measurement_names"]
  t <- t[t!="ID.Patient"]
  t <- t[t!="WellPosition"]
  t <- t[!t %in% c("File.Name", "Creation.Date", "Batch", "Class", "Patient_ID", "batch_qc")]
  t <- t[!t %in% names(colData(mae))]
})
tech_vars_continuous
```

**NOTE:** consider add sum (total carnitines) and ratios (vlcfa: C24/C22 and C26/C22)

Phenotype information shared across all assays can be accessed via `colData()`

```{r}
names(colData(mae))
```


```{r}
# variable describing both Sex and PhenotypicFeature
colData(mae)$Sex.PhenotypicFeature <- gsub(
  ".NA", "", paste(colData(mae)$Sex, colData(mae)$PhenotypicFeature, sep = "."))
```



## Principal component analysis -- PCA

PCA performed on standardized (centered to mean and scaled to unit-variance) data.

```{r run_pca}
run_pca_se <- function(se) {
  # run PCA on SummarizedExperiment
  a <- t(assays(se)[[1]])
  # remove non-variable features
  a <- a[, apply(a, 2, function(column) {
    var(column) != 0 })]
  # standardize and run PCA
  pca <- prcomp(a, center = T, scale. = T)
}

pca <- lapply(experiment_names, function(n) {run_pca_se(mae[[n]])})
```


### PCA -- scree plots 

Scree plots are bar plots showing the explained variation per Principal Component (PC).
Cumulative explained variation is shown as a line.

```{r pcs_scree_plots}
library(ggplot2)
#'  Create a screeplot of the variance explained by principal components
#'  
#'  @param pc PCA as input
#'  @param name_file Name of the barplot (.png file)
#'  @return The screeplot object 
make_screeplot <- function(pc, main_title) {
  require(ggplot2)
  require(ggbreak)
  # create summary of pca
  vars <- pc$sdev^2
  vars <- vars/sum(vars)
  df <- as.data.frame(cbind(
    'SD' = pc$sdev, 'Prop' = vars, 'cumulative' = cumsum(vars)))
  df$PC <- seq(nrow(df))
  df$Prop <- df$Prop * 100
  df$cumulative <- df$cumulative * 100
  df <- df[c('PC', 'SD', 'Prop', 'cumulative')]
  df <- as.data.frame(lapply(df, unlist))
  
  # Get number of PCs that explain 80%
  eighty_percent <-  min(which(df$cumulative > 80))
  scale <- round(eighty_percent/25)
  if (scale == 0) {
    scale <- 1
  }
  
  # Screeplot
  plot1 <- ggplot(data = df[df$PC <= eighty_percent+3, ], 
                  aes(x = PC, y = Prop)) + #reorder(PC, -Prop)
    geom_bar(stat = 'identity') +
    geom_line(aes(y=cumulative,group=1 ), size = 1, color = 'red') +
    xlab('Principal Component') +
    ylab('Proportion of variance explained (%)') +
    theme(plot.title = element_text(size = 12, hjust = 0.5),
          legend.position = 'none') +
    ggtitle(main_title)
  plot1 <- plot1 + 
    geom_vline(aes(xintercept = eighty_percent), size = 2) +
    geom_label(aes(x = eighty_percent, y = 0.50,
                   label = '80%', vjust = -1, size = 8)) 
  plot1 <- plot1 +
    scale_x_continuous(breaks = seq(1, length(df$PC), by = 1), position = "bottom")
  if (eighty_percent > 18) {
    plot1 <- plot1 + scale_x_break(breaks = c(13, eighty_percent-3))
  }
  return(plot1)
}


# lapply(names(pca), function(x) {
#   make_screeplot(pca[[x]], paste0("PCA, scree plot\n", x))
#})
```


### PCA -- score plots 

#### Pairwise scatter plots of PC scores colored by biological and technical variables

```{r pca_score_plots}
library(ggplot2)
library(GGally)
library(ggokabeito) # Okabe-Ito

bio_vars <- c(
  "Sex",
  "Sex.PhenotypicFeature", 
  "Age.group",
  "BMI.group",
  "Smoking.Status",
  "ABO.Blood.Group",
  "Rh.Blood.Group")

num_PCs <- 3

lapply(names(pca), function(assay_name) { 
  lapply(c(bio_vars, tech_vars_factor[[assay_name]]), function(color_var) {
    message("PCA score plots - ", assay_name, ", ", color_var)
    # PCA scores 
    pca_scores <- as.data.frame(pca[[assay_name]]$x[, 1:num_PCs])
    # data for coloring
    if (color_var %in% bio_vars) {
      pca_scores[[color_var]] = colData(mae)[rownames(pca_scores), color_var]
    } else {
      pca_scores[[color_var]] = colData(experiments(mae)[[assay_name]])[
        rownames(pca_scores), color_var]
    }
    num_factor_levels <- length(levels(as.factor(pca_scores[[color_var]])))
    if (num_factor_levels > 1) { # multiple factor levels
      # refactor NA to "NA" - otherwise not shown in ggpairs plot
      levels(pca_scores[[color_var]]) <- c(levels(pca_scores[[color_var]]), "NA")
      pca_scores[[color_var]][is.na(pca_scores[[color_var]])] <- "NA"
      # pairs plot of PCs
      ggp <- ggpairs(
        pca_scores, 
        columns = 1:num_PCs,
        aes(color = .data[[color_var]]),#, shape = .data[[color_var]], alpha = 0.25),
        progress = FALSE,
        diag = list(continuous = "barDiag"),
        upper = list(continuous = "blank"),
        legend = 1
      ) +
        ggtitle(stringr::str_wrap(paste0(
          "PCA score plot, color by ", color_var, ", ", assay_name), width = 60)) +
        theme_bw() +
        # suppress showing alpha in legend
        scale_alpha(guide = "none")
      # change color palette
      for (i in 1:ggp$nrow) {
        for (j in 1:ggp$ncol) {
          ggp[i, j] <- ggp[i, j] +
            # color blind friendly palette: Okabe-Ito
            scale_color_okabe_ito() + scale_fill_okabe_ito()
        }
      }
      ggp + theme(legend.position = "bottom") 
    } else {
      ggp <- NULL
    }
    ggp
  })
})
```

#### Box plots of PC scores colored by biological and technical variables

Outlier samples are labeled. Statistical tests are performed to determine if PCs are significantly associated with biological or technical variables

```{r pca_box_plots, results='asis'}
library(reshape2)
library(dplyr)
library(ggrepel)
library(ggpubr)
num_PCs <- 10

detect_outlier <- function(x) {
  {x < quantile(x, 0.25) - 1.5 * IQR(x)} | {
    x > quantile(x, 0.75) + 1.5 * IQR(x)}
}

shorten_name <- function(filename) {
  filename <- gsub(" ", "", filename, fixed = T)
  filename <- gsub("Verylongchainfattyacids", "VLCFA", filename, fixed = T)
  filename <- gsub("imputedmissingvalues", "imputed", filename, fixed = T)
  filename <- gsub("cell-typeadjusted", "celltypeadj", filename, fixed = T)
  filename <- gsub("|", "_", filename, fixed = T)
  filename <- gsub("batch-adjusted", "batchadj", filename, fixed = T)
  filename <- gsub("100,000", "100000", filename, fixed = T)
  filename <- gsub("CpGsites", "CpGs", filename, fixed = T)
  filename
}

box_plots <- list()
significant_associations_factor <- list()
i <- 1

for (assay_name in names(pca)) {
  for (color_var in c(bio_vars, tech_vars_factor[[assay_name]])) {
    message("PCA scores box plots - ", assay_name, ", ", color_var)
    # PCA scores 
    num_PCs_ <- ifelse(num_PCs > ncol(pca[[assay_name]]$x),
                       ncol(pca[[assay_name]]$x), num_PCs)
    pca_scores <- as.data.frame(pca[[assay_name]]$x[, 1:num_PCs_])
    # data for coloring
    if (color_var %in% bio_vars) {
      pca_scores[[color_var]] = as.factor(
        colData(mae)[rownames(pca_scores), color_var])
    } else {
      pca_scores[[color_var]] = as.factor(
        colData(experiments(mae)[[assay_name]])[rownames(pca_scores), color_var])
    }
    if (length(levels(pca_scores[[color_var]])) > 1) { # multiple factor levels
      pca_scores$subject.id <- rownames(pca_scores)
      pca_scores_long <- pca_scores %>%
        melt(id.vars = c("subject.id", color_var)) %>%
        group_by(across(c(all_of(color_var), "variable"))) %>%
        mutate(outlier = ifelse(detect_outlier(value), 
                                gsub("CZC", "", subject.id), NA))
      # compare mean values of groups - significance tests
      stat_tests <- compare_means(
        as.formula(paste0('value~', color_var)),
        data = pca_scores_long, group.by = "variable")
      stat_tests$assay_name <- assay_name
      stat_tests$tested_variable <- color_var
      df <- data.frame(stat_tests[stat_tests$p < 0.05, ])
      # if (nrow(df) > 0) {
      significant_associations_factor[[i]] <- local(df)
      # }
      print(knitr::kable(stat_tests))
      cat("\n\n") # some extra lines for kable
      # box plot
      ggp <- ggplot(pca_scores_long, 
             aes(x = variable, y = value, fill = .data[[color_var]])) +
        geom_boxplot(position = position_dodge(width = 0.9)) +
        geom_text_repel( # label outliers
          aes(label = outlier), na.rm = TRUE, 
          position = position_dodge(width = 0.9)) + 
        scale_fill_okabe_ito() + 
        #stat_compare_means() +
        ggtitle(stringr::str_wrap(paste0(
          "Box plot, PC scores, color by ", color_var, ", ", assay_name), 
          width = 60)) +
        theme_bw() +
        theme(
          axis.title.x = element_blank(),
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          # panel.spacing.x = unit(0, "lines"),
          legend.position = "bottom") +
        facet_wrap(vars(variable), nrow = 1, scales = "free") 
      filename <- shorten_name(
        paste0("PCAscoreboxplot_", assay_name, "_", color_var, ".png"))
      ggsave(file.path(outdir, filename), ggp,
             width = 28.04, height = 10.46, units = "cm", dpi = 300)
      box_plots[[i]] <- local(ggp)
      i <- i + 1
    }
  }
}
```


```{r, fig.width=20}
lapply(box_plots, plot)
```

```{r}
significant_associations_factor <- do.call(rbind, significant_associations_factor)
head(significant_associations_factor)
dim(significant_associations_factor)
dim(significant_associations_factor[significant_associations_factor$p.adj<0.05,])
```


### Correlation of PCs with biological variables

```{r pca_corr_plots}
library(corrplot)

bio_vars_continuous <- c(
  "Age",
  "Height",
  "Weight",
  "BMI",
  "Leukocytes",
  "Erythrocytes",
  "Hemoglobin",
  "Hematocrit",
  "Platelets",
  "MCV",
  "MCH",
  "MCHC",
  "Lymphocytes",
  "MXD",
  "Neutrophils")

sig.level <- 0.05

significant_associations_contin <- lapply(names(pca), function(assay_name) {
  message("Correlations - ", assay_name)
  num_PCs_calculated <- dim(pca[[assay_name]]$x)[2]
  print(num_PCs_calculated)
  expl_var <- pca[[assay_name]]$sdev^2 / sum(pca[[assay_name]]$sdev^2)
  num_PCs_todisplay <- ifelse(num_PCs > num_PCs_calculated, 
                              num_PCs_calculated, num_PCs)
    #sum(expl_var>0.02)
  
    
  expl_var <- expl_var[1:num_PCs_todisplay]
  
  pca_scores <- as.data.frame(
    pca[[assay_name]]$x[, 1:num_PCs_todisplay])
  phenotype_continuous = as.data.frame(
    colData(mae)[rownames(pca_scores), bio_vars_continuous])
  phenotype_continuous <- cbind(
    phenotype_continuous,
    subset(data.frame(colData(mae[[assay_name]]))[rownames(pca_scores), ],
           select = tech_vars_continuous[[assay_name]]))
  # calculate correlation
  score_pheno_cor <- cor(
   pca_scores, phenotype_continuous, 
   method = "pearson", use = "pairwise.complete.obs")
  # significance test
  sig_test <- psych::corr.test(
    pca_scores, phenotype_continuous,
    use = "pairwise",
    method = "pearson",
    adjust = "BH",
    alpha = sig.level
  )
  
  significant_cor <- NULL
  if (sum(sig_test$p < 0.05) > 0) {
    significant_cor <- data.frame(sig_test$p)
    significant_cor$variable <- rownames(significant_cor)
    significant_cor <- tidyr::gather(
      significant_cor, "tested_variable", "p", -variable)
    significant_cor <- significant_cor[significant_cor$p < 0.05, ]
    significant_cor$p.adj <- apply(significant_cor, 1, function(row) {
      sig_test$p.adj[row[["variable"]], row[["tested_variable"]]]
    })
    significant_cor$assay_name <- assay_name
    significant_cor$method <- "Pearson"
    significant_cor$.y. <- significant_cor$group1 <- significant_cor$group2 <-
      significant_cor$p.format <- significant_cor$p.signif <- NA
  }
  
  rownames(score_pheno_cor) <- paste0(
    sprintf("%-4s", rownames(score_pheno_cor)), 
    " ",  
    sprintf("%3s", scales::percent(
      round(expl_var,2))))
  
  # correlation plot with FDR-adjusted p-values 
  cor_lim <- max(abs(c(min(score_pheno_cor), max(score_pheno_cor))))
  
  png(file.path(outdir, 
                paste0("PCA_cor_", shorten_name(assay_name), ".png")),
    width = 28.04, height = 10.46, units = "cm", res = 300)
  corrplot(
    score_pheno_cor, 
    is.corr = FALSE,
    col.lim = c(-cor_lim, cor_lim),
    title = stringr::str_wrap(assay_name, width = 30),
    mar = c(0, 0, 3, 0),
    tl.col = "black",
    p.mat = sig_test$p.adj,
    sig.level = sig.level)
  dev.off()
  corrplot(
    score_pheno_cor, 
    is.corr = FALSE,
    col.lim = c(-cor_lim, cor_lim),
    title = stringr::str_wrap(assay_name, width = 30),
    mar = c(0, 0, 3, 0),
    tl.col = "black",
    p.mat = sig_test$p.adj,
    sig.level = sig.level)
  
  # make scatter plot of PCs and variables with significant correlation
  sig_pairs <- as.data.frame(which(sig_test$p.adj < sig.level, arr.ind = TRUE))
  sig_pairs$p.adj <- sig_test$p.adj[which(sig_test$p.adj < sig.level)]
  sig_pairs <- sig_pairs[order(sig_pairs$row), ] # sort by PCs
  if (dim(sig_pairs)[1] > 0) {
    apply(sig_pairs, 1, function(x) {
      pc_ <- rownames(sig_test$p.adj)[[x[["row"]]]]
      bio_ <- colnames(sig_test$p.adj)[[x[["col"]]]]
      df <- data.frame(pc_ = pca_scores[, pc_], 
                       bio_ = phenotype_continuous[, bio_])
      names(df) <- c(pc_, bio_)
      ggp <- ggplot(df, aes(x = .data[[pc_]], y = .data[[bio_]])) +
        geom_point() +
        stat_smooth(method = "lm") +
        stat_cor(method = "pearson") +
        ggtitle(stringr::str_wrap(assay_name, width = 60)) +
        theme_bw()
      plot(ggp)
      NULL
    })
  }
  significant_cor
})
```


```{r}
significant_associations_contin <- do.call(rbind, significant_associations_contin)
head(significant_associations_contin)
```


# Change to readable (shorter) names
```{r}
assay_names <- config$experiment_names$PCA

# Mapping of original to readable names
assay_name_map <- config$readable_names

# Translate assay names using the map
shorter_assay_names <- sapply(assay_names, function(x) {
  if (!is.null(assay_name_map[[x]])) {
    assay_name_map[[x]]
  } else {
    x  # fallback to original name
  }
})


# Create a data frame for reference
assay_names_df <- data.frame(
  original = assay_names,
  readable = shorter_assay_names,
  stringsAsFactors = FALSE
)
assay_names_df
```


```{r}
get_shorter_assay_names <- function(assay_name) {
  
  if (grepl(pattern = 'Lipidomics, negative', x = assay_name)){
     short_name <- 'Lipidomics Set 1 (N)'
  }
  else if (grepl(pattern = 'Lipidomics, positive', x = assay_name)){
     short_name <- 'Lipidomics Set 1 (P)'
  }
  else if (grepl(pattern = 'Lipidomics-SPLINE', x = assay_name)){
     short_name <- 'Lipidomics Set 2'
  }
  else if (grepl(pattern = 'Very long chain', x = assay_name)){
     short_name <- 'VLCFA'
  }
  else if (grepl(pattern = 'Acylcarnitines', x = assay_name)){
     short_name <- 'Acylcarnitines'
  }
  else if (grepl(pattern = 'Amino acids', x = assay_name)){
     short_name <- 'Amino acids'
  }
  else if (grepl(pattern = 'IPP', x = assay_name)){
     short_name <- 'Proteomics-IPP'
  }
  else if (grepl(pattern = 'Proteomics', x = assay_name)){
     short_name <- 'Proteomics'
  }
  else if (grepl(pattern = 'mRNA-seq', x = assay_name)){
     short_name <- 'mRNA-seq'
  }
  else if (grepl(pattern = 'miRNA-seq', x = assay_name)){
     short_name <- 'miRNA-seq'
  }
  else if (grepl(pattern = 'EM-seq', x = assay_name)){
     short_name <- 'EM-seq'
  }
  else{
    short_name <- assay_name
  }
  
  short_name
}  

# Retrieve list of shorter assay names
shorter_assay_names <- lapply(experiment_names, get_shorter_assay_names)

assay_names_temp <- experiment_names

assay_names_df <- data.frame(assay_names_temp, as.character(shorter_assay_names))
#names(assay_names) <- shorter_assay_names

assay_names_df
```

```{r, fig.width=10, fig.height=5}
significant_associations <- rbind(
  significant_associations_factor, 
  significant_associations_contin[, names(significant_associations_factor)])

significant_associations <- significant_associations[
  significant_associations$tested_variable != "Sex.PhenotypicFeature",]
if ("Group" %in% significant_associations$tested_variable) {
  significant_associations[
    significant_associations$tested_variable == "Group", 
    ]$tested_variable <- "Batch"
}
if ("Measurement.date" %in% significant_associations$tested_variable) {
  significant_associations[
    significant_associations$tested_variable == "Measurement.date",
    ]$tested_variable <- "Batch"
}

unique(significant_associations$tested_variable)
significant_associations$tested_variable <- factor(
  significant_associations$tested_variable,
  levels = c(
    "Sex", 
    "Height", "Weight", "BMI", "BMI.group", 
    "Age", "Age.group", "Smoking.Status", 
    "ABO.Blood.Group", "Rh.Blood.Group", 
    "Leukocytes", 
    "Neutrophils", 
    "Lymphocytes", 
    "MXD", 
    "Erythrocytes", 
    "Hemoglobin", 
    "Hematocrit", 
    "MCHC", 
    "MCV", 
    "MCH", 
    "Platelets", 
    "Injection_Order", "Batch"))

unique(significant_associations$assay_name)
significant_associations$assay_name <- factor(
  significant_associations$assay_name,
  levels = experiment_names
)

significant_associations$expl_var <- apply(
  significant_associations, 1, function(row) {
    vars <- pca[[row[["assay_name"]]]]$sdev^2
    vars <- vars/sum(vars)
    vars[as.numeric(gsub("PC", "", row[["variable"]]))]
  })


library(dplyr)
data2plot <- significant_associations %>%
  filter(p.adj < 0.05) %>%
  group_by(assay_name, tested_variable) %>%
  summarise(`Cumulative variation\nexplained by significant PCs` = sum(expl_var))


data2plot <- data2plot[!is.na(data2plot$tested_variable), ] # Remove rows with NA value for tested_variable.
#TODO: Add miRNA-qPCR_RT continuous tech variable to bio_vars_continuous. Or technical vars?

data2plot$`Variable type` <- "Clinical hematology"
data2plot[data2plot$tested_variable == "Batch", ]$`Variable type` <- "Technical"
data2plot[data2plot$tested_variable %in% c(
  "Sex", "Height", "Weight", "BMI", "BMI.group", 
  "Age", "Age.group", "Smoking.Status", "ABO.Blood.Group", "Rh.Blood.Group",
  "Age_pred_BLUP", "Age_acc_BLUP", "Age_pred_Levine", "Age_acc_Levine"), 
  ]$`Variable type` <- "Phenotype"


# Add shorter names
shorter_names <- lapply(data2plot$assay_name, function(x) {
    new_name <- assay_names_df[assay_names_df$assay_names_temp == x, 2]
    new_name
}) 
data2plot$assay_name <- shorter_names
data2plot$assay_name <- as.factor(as.character(data2plot$assay_name))

# Manually add missing associations, to include in the plot anyway
temp_df <- data.frame(temp='miRNA-seq', temp0='Age', temp1=NA, temp2='Phenotype')
temp_df <- rbind(temp_df, c(temp='miRNA-seq', temp0='Age', temp1=NA, temp2='Phenotype'))
temp_df$temp1 <- as.numeric(temp_df$temp1)
colnames(temp_df) <- colnames(data2plot)
data2plot <- rbind(data2plot, temp_df)

ggp <- ggplot(data2plot) +
  geom_point(aes(x = factor(.data$tested_variable, levels = c(
    "Sex", 
    "Height", "Weight", "BMI", "BMI.group", 
    "Age", "Age.group", "Smoking.Status", 
    "ABO.Blood.Group", "Rh.Blood.Group", 
    "Leukocytes", 
    "Neutrophils", 
    "Lymphocytes", 
    "MXD", 
    "Erythrocytes", 
    "Hemoglobin", 
    "Hematocrit", 
    "MCHC", 
    "MCV", 
    "MCH", 
    "Platelets", 
    "Injection_Order", "Batch")),
                 y = .data$assay_name,
                 size = .data$`Cumulative variation\nexplained by significant PCs`,
                 col = .data$`Variable type`)) +
  theme(
    axis.text.x = element_text(angle = 30, vjust = 1, hjust = 1)
  ) +
  scale_color_okabe_ito() +
  ylab("Data set") +
  xlab("Biological/technical variable") +
  ggtitle("Significant associations between top 10\nPrincipal Components and variables")

ggsave(
  file.path(outdir, "PCA_summary.pdf"),
  plot = ggp,
  width = 23, height = 10.46, units = "cm", dpi = 500)

ggp
```



## Biological vs Chronological Age


```{r, fig.width=10, fig.height=5}
ggp <- ggplot(data2plot) +
  geom_point(aes(x = factor(.data$tested_variable, levels = c(
    "Age", "Age.group",
    "Age_pred_BLUP", "Age_acc_BLUP",
    "Age_pred_Levine", "Age_acc_Levine")),
                 y = .data$assay_name,
                 size = .data$`Joint variation\nexplained by PCs`,
                 col = .data$`Variable type`)) +
  theme(
    axis.text.x = element_text(angle = 30, vjust = 1, hjust = 1)
  ) +
  scale_color_okabe_ito() +
  ylab("Data set") +
  xlab("Biological/technical variable") +
  ggtitle("Significant associations between top 10\nPrincipal Components and variables")

ggsave(
  file.path(outdir, "PCA_summary_Age.png"),
  plot = ggp,
  width = 23, height = 10.46, units = "cm", dpi = 300)

ggp
```

## Session information

```{r}
sessionInfo()
```


