---
title: 'Multi-block PLS-DA (DIABLO)'
author:
- Anna Niehues^[Radboud university medical center, Anna.Niehues@radboudumc.nl]
output:
  html_document:
    toc: true
    df_print: paged
bibliography: references.bib
params:
  outcome: "Sex"
  n_comp: 100
  n_repeat: 0
  config_file: NULL
  references: NULL
  mae_hdf5_dir_path: NULL
---

Perform N-integration discriminant analysis with DIABLO implemented in mixOmics. 

**NOTE:** set n_repeat to a high enough value for model training and evaluation (50-100)

```{r}
if (paste0(R.Version()$major, ".", R.Version()$minor) != "4.2.1") {
  warning("This R notebook was implemented using R version 4.2.1")
}
print(params)
config <- yaml::read_yaml(params$config)
outdir <- config$output_directory$DIABLO
dir.create(outdir)
```


## Read input data 

Input data is a MultiAssayExperiment (MAE) object. All assays in the MAE will be used for analysis Multi-omics Factor Analysis (MOFA).

```{r}
library(MultiAssayExperiment)
library(HDF5Array)

mae <- loadHDF5MultiAssayExperiment(dir = config$mae_hdf5_dir_path, 
                                    prefix = config$mae_hdf5_prefix)
```

Select male/female individuals based on input parameters

```{r}
sample_selection <- c()
if (config$sample_selection$include_MALE) {
  sample_selection <- c(sample_selection, "MALE")}
if (config$sample_selection$include_FEMALE) {
  sample_selection <- c(sample_selection, "FEMALE")}

mae <- mae[, colData(mae)$Sex %in% sample_selection]

mae
```



```{r}
# experiments(mae)
assay_names <- config$experiment_names$DIABLO
names(assay_names) <- assay_names
assay_names
```


## Prepare data for DIABLO

* X - named list of data sets (blocks) measured on the same samples (matrices, samples x variables, with samples order matching in all blocks
* Y - vector with classes for observations

```{r}
sample_names <- Reduce(
  intersect,
  lapply(assay_names, function(assay_name) {
    rownames(colData(mae[[assay_name]]))
}))

features2include <- lapply(assay_names, function(assay_name) {
  rownames(assays(mae)[[assay_name]])[
    apply(as.data.frame(assays(mae)[[assay_name]]), 1, function(row) {
      var(row) != 0 & sum(row == 0) < 10 })]
      # var(row[sample_names]) != 0 & sum(row[sample_names] == 0) < 10 })]
}) 
lapply(features2include, length)


# X <- lapply(assay_names, function(assay_name) {
#   t(assays(mae)[[assay_name]])[, features2include[[assay_name]]]
# })
X <- lapply(assay_names, function(assay_name) {
  t(assays(mae)[[assay_name]])[sample_names, features2include[[assay_name]]]
})
Y <- colData(mae)[sample_names, ][[params$outcome]]

# remove samples with missing outcome 
X <- lapply(X, function(x) {
  x <- x[!is.na(Y), ]
})
Y <- Y[!is.na(Y)]

if (is.numeric(Y)) {
  # make categories
  Y <- cut(Y, breaks = quantile(Y, probs = seq(0,1,1/3)), include.lowest = T)
}

print(lapply(X, dim))
print(summary(Y))
```



## Pairwise PLS


```{r}
list.keepX <- list.keepY <- lapply(assay_names, function(assay_name) {
  ifelse(ncol(X[[assay_name]]) > 1000, 1000, ncol(X[[assay_name]]))
})

combos <- combn(assay_names, 2, simplify = FALSE)

pairwise.pls.cor <- lapply(combos, function(combo) {
  #print(combo)
  samples_names <- intersect(rownames(X[[combo[[1]]]]), rownames(X[[combo[[2]]]]))
  pls.XY <- spls(
    X[[combo[[1]]]][samples_names,], X[[combo[[2]]]][samples_names,], 
    keepX = list.keepX[[combo[[1]]]], 
    keepY = list.keepY[[combo[[2]]]],
    mode = "canonical")
  cat("\n")
  
  plotVar(
    pls.XY, #cutoff = 0.4,
    title = paste0("sPLS, ", combo[[1]], ", ", combo[[2]]),
    legend = combo, var.names = F, pch = c(16, 17))

  list(
    combo = combo,
    cor.combo = cor(pls.XY$variates$X[, "comp1"], pls.XY$variates$Y[, "comp1"])
  )
})

cor.combos.sum <- sum(sapply(pairwise.pls.cor, function(x) {x[["cor.combo"]]}))

design <- matrix(0, ncol = length(assay_names), nrow = length(assay_names),
                 dimnames = list(assay_names, assay_names))

for (x in pairwise.pls.cor) {
  cor_val <- round(x[["cor.combo"]] / cor.combos.sum / 2, 2)
  design[x[["combo"]][[1]], x[["combo"]][[2]]] <- cor_val
  design[x[["combo"]][[2]], x[["combo"]][[1]]] <- cor_val
}

cat("Design matrix for DIABLO\n")
design
```


## Get omics names list
```{r}
assay_names <- config$experiment_names$DIABLO

# Mapping of original to readable names
assay_name_map <- config$readable_names

# Translate assay names using the map
shorter_assay_names <- sapply(assay_names, function(x) {
  if (!is.null(assay_name_map[[x]])) {
    assay_name_map[[x]]
  } else {
    x  # fallback to original name
  }
})


# Create a data frame for reference
assay_names_df <- data.frame(
  original = assay_names,
  readable = shorter_assay_names,
  stringsAsFactors = FALSE
)

print(assay_names_df)
```


```{r, fig.height=8}
library(ggcorrplot)

pairwise.pls.cor_m <- matrix(
  NA, ncol = length(assay_names), nrow = length(assay_names),
  dimnames = list(assay_names, assay_names))

for (x in pairwise.pls.cor) {
  print(x)
  cor_val <- x[["cor.combo"]]
  pairwise.pls.cor_m[x[["combo"]][[1]], x[["combo"]][[2]]] <- cor_val
  pairwise.pls.cor_m[x[["combo"]][[2]], x[["combo"]][[1]]] <- cor_val
}

# Shorter omics names
rownames(pairwise.pls.cor_m) <- assay_names_df$readable
colnames(pairwise.pls.cor_m) <- assay_names_df$readable


dimnames(pairwise.pls.cor_m) <- lapply(
  dimnames(pairwise.pls.cor_m), function(x) {
    stringr::str_wrap(x, width = 35) })

pairwise.pls.cor_m[is.na(pairwise.pls.cor_m)] <- 1


ggp <- ggcorrplot(
  data.frame(pairwise.pls.cor_m),
  lab = T,
  hc.order = T) +
  scale_fill_gradient2(
  limit = c(0.3, 1), 
  low = "white", high = "blue")

ggsave(file.path(outdir, "pairwise_sPLS_cor.png"), ggp,
       width = 22, height = 20, units = "cm", dpi = 300)
ggp
```


## R session information

```{r}
sessionInfo()
```

## References

