---
title: "Create MultiAssayExperiment object"
output:
  html_document:
    df_print: paged
bibliography: references.bib
params:
  phenotype_data: Z:/omics/Phenotype/phenotypes_from_phenopackets_2022-12-01.csv
  
  acylcarnitines_data: Z:/omics/MetabolomicsTargeted_RUMC/data_for_integration/woQC_flt_ln_std/metabolomics_acylcarnitines_rumc_FIA-MS_positive__metabolite_profiling.csv
  acylcarnitines_data_unscaled: Z:/omics/MetabolomicsTargeted_RUMC/data_for_integration/woQC_flt_ln/metabolomics_acylcarnitines_rumc_FIA-MS_positive__metabolite_profiling.csv
  acylcarnitines_data_batchadjusted: Z:/omics/MetabolomicsTargeted_RUMC/data_for_integration/woQC_flt_ln_ComBatCovSex_std/metabolomics_acylcarnitines_rumc_FIA-MS_positive__metabolite_profiling.csv
  acylcarnitines_data_batchadjusted_imputed: Z:/omics/MetabolomicsTargeted_RUMC/data_for_integration/woQC_flt_ln_ComBatCovSex_impLAR_std/metabolomics_acylcarnitines_rumc_FIA-MS_positive__metabolite_profiling.csv
  acylcarnitines_sample_metadata: Z:/omics/MetabolomicsTargeted_RUMC/data_for_integration/technical_parameters/metabolomics_acylcarnitines_rumc_FIA-MS_positive__metabolite_profiling.csv
  acylcarnitines_feature_metadata: Z:/omics/MetabolomicsTargeted_RUMC/ISA_and_MAF_files_2023-02-15/m_metabolomics_acylcarnitines_rumc_FIA-MS_positive__metabolite_profiling_v2_maf.tsv
  
  aminoacids_data: Z:/omics/MetabolomicsTargeted_RUMC/data_for_integration/woQC_flt_ln_std/metabolomics_aminoacids_mumc_LC-MS_positive_reverse-phase_metabolite_profiling.csv
  aminoacids_data_unscaled: Z:/omics/MetabolomicsTargeted_RUMC/data_for_integration/woQC_flt_ln/metabolomics_aminoacids_mumc_LC-MS_positive_reverse-phase_metabolite_profiling.csv
  aminoacids_data_batchadjusted: Z:/omics/MetabolomicsTargeted_RUMC/data_for_integration/woQC_flt_ln_ComBatCovSex_std/metabolomics_aminoacids_mumc_LC-MS_positive_reverse-phase_metabolite_profiling.csv
  aminoacids_data_batchadjusted_imputed: Z:/omics/MetabolomicsTargeted_RUMC/data_for_integration/woQC_flt_ln_ComBatCovSex_impLAR_std/metabolomics_aminoacids_mumc_LC-MS_positive_reverse-phase_metabolite_profiling.csv
  aminoacids_sample_metadata: Z:/omics/MetabolomicsTargeted_RUMC/data_for_integration/technical_parameters/metabolomics_aminoacids_mumc_LC-MS_positive_reverse-phase_metabolite_profiling.csv
  aminoacids_feature_metadata: Z:/omics/MetabolomicsTargeted_RUMC/ISA_and_MAF_files_2023-02-15/m_metabolomics_aminoacids_mumc_LC-MS_positive_reverse-phase_metabolite_profiling_v2_maf.tsv
  
  fattyacids_data: Z:/omics/MetabolomicsTargeted_RUMC/data_for_integration/woQC_flt_ln_std/metabolomics_fattyacids_mumc_LC-MS_negative_reverse-phase_metabolite_profiling.csv
  fattyacids_data_unscaled: Z:/omics/MetabolomicsTargeted_RUMC/data_for_integration/woQC_flt_ln_std/metabolomics_fattyacids_mumc_LC-MS_negative_reverse-phase_metabolite_profiling.csv
  fattyacids_data_batchadjusted: Z:/omics/MetabolomicsTargeted_RUMC/data_for_integration/woQC_flt_ln_ComBatCovSex_std/metabolomics_fattyacids_mumc_LC-MS_negative_reverse-phase_metabolite_profiling.csv
  fattyacids_sample_metadata: Z:/omics/MetabolomicsTargeted_RUMC/data_for_integration/technical_parameters/metabolomics_fattyacids_mumc_LC-MS_negative_reverse-phase_metabolite_profiling.csv
  fattyacids_feature_metadata: Z:/omics/MetabolomicsTargeted_RUMC/ISA_and_MAF_files_2023-02-15/m_metabolomics_fattyacids_mumc_LC-MS_negative_reverse-phase_metabolite_profiling_v2_maf.tsv
  
  lipidomics_neg_data: Z:/omics/LipidomicsUntargeted_IMTM/Untargeted_Lipidomics_IMTM_LN_PutativelyAnnotatedOnly_NEG_EATRIS_20220902_DataArea_Log.txt
  lipidomics_neg_sample_metadata: Z:/omics/LipidomicsUntargeted_IMTM/Untargeted_Lipidomics_IMTM_LN_PutativelyAnnotatedOnly_NEG_EATRIS_20220902_MetaDataSamples.txt
  lipidomics_neg_feature_metadata: Z:/omics/LipidomicsUntargeted_IMTM/Untargeted_Lipidomics_IMTM_LN_PutativelyAnnotatedOnly_NEG_EATRIS_20220902_MetaDataFeature.txt
  lipidomics_pos_data: Z:/omics/LipidomicsUntargeted_IMTM/Untargeted_Lipidomics_IMTM_LN_PutativelyAnnotatedOnly_POS_EATRIS_20220902_DataArea_Log.txt
  lipidomics_pos_sample_metadata: Z:/omics/LipidomicsUntargeted_IMTM/Untargeted_Lipidomics_IMTM_LN_PutativelyAnnotatedOnly_POS_EATRIS_20220902_MetaDataSamples.txt
  lipidomics_pos_feature_metadata: Z:/omics/LipidomicsUntargeted_IMTM/Untargeted_Lipidomics_IMTM_LN_PutativelyAnnotatedOnly_POS_EATRIS_20220902_MetaDataFeature.txt
  
  lipidomics_spline_pos_data: Z:/omics/LipidomicsUntargeted_IMTM/New_Data_2025/processed_data_EatrisSamples.csv
  lipidomics_spline_sample_metadata: Z:/omics/LipidomicsUntargeted_IMTM/New_Data_2025/sample_metadata_EatrisSamples.csv
  lipidomics_spline_feature_metadata: Z:/omics/LipidomicsUntargeted_IMTM/New_Data_2025/variable_metadata.csv
  
  proteomics_data: Z:/omics/Proteomics_IMTM/Upload_2023-03-03/proteins_multiomics_batchCorrected-ComBat-sex.txt
  proteomics_data_unscaled: 
  proteomics_data_imputed: Z:/omics/Proteomics_IMTM/Upload_2023-03-03/proteins_multiomics_batchCorrected-ComBat-sex_imputed.txt
  proteomics_sample_metadata: Z:/omics/Proteomics_IMTM/Upload_2023-02-02/samples_annotation.txt
  proteomics_feature_metadata: Z:/omics/Proteomics_IMTM/Upload_2023-02-02/proteins_annotation.txt
  
  proteomics_ipp_data: Z:/omics/Proteomics_IMTM/Upload_2025-05-22_IPP/IPP_log10_batchEffectRemoved_std.txt
  proteomics_ipp_feature_metadata: Z:/omics/Proteomics_IMTM/Upload_2025-05-22_IPP/IPP_featureAnnotation.txt
  proteomics_ipp_sample_metadata: Z:/omics/Proteomics_IMTM/Upload_2025-05-22_IPP/IPP_batchInfo.txt
  
  mRNAqRTPCR_data: Z:/omics/microRNAqRTPCR_SERMAS/Data_Integration/qPCR_miRNA_to_Integration.tsv
  mRNAqRTPCR_sample_metadata: Z:/omics/microRNAqRTPCR_SERMAS/Analysis/QC_Hemolisis_data.tsv
  
  aCGH_data: Z:/omics/aCGH_IMTM/multi-omics-ready/aCGH_geneTable_aberrated_imputed_10-2025_116samples.csv
  aCGH_sample_metadata: Z:/omics/aCGH_IMTM/multi-omics-ready/batch.txt
  
  miRNAseq_mature_data: Z:/omics/microRNAseq_FIMM/miRBase_mature/mature_normalized_CPM.txt
  miRNAseq_mature_data_batchadjusted: Z:/omics/microRNAseq_FIMM/miRBase_mature/mature_Log_Combat_CovSex.csv
  miRNAseq_mature_data_batchadjusted_celltypeadjusted: Z:/omics/microRNAseq_FIMM/miRBase_mature/mature_Log_Combat_CovSex_corrected.csv
  miRNAseq_hairpin_data: Z:/omics/microRNAseq_FIMM/miRBase_hairpin/hairpin_normalized_CPM.txt
  miRNAseq_sample_metadata: Z:/omics/microRNAseq_FIMM/metadata.csv
  
  mRNAseq_data: Z:/omics/mRNAseq_FIMM/normalized-cpm-count.tsv
  mRNAseq_data_batchadjusted: Z:/omics/mRNAseq_FIMM/normalized-cpm-count_Combat_CovSex.csv
  mRNAseq_data_batchadjusted_corrected: Z:/omics/mRNAseq_FIMM/normalized-cpm-count_Combat_CovSex_Log_corrected.csv
  mRNAseq_data2_batchadjusted_vst: Z:/omics/mRNAseq_FIMM/raw_counts_Deseq_vst.csv
  mRNAseq_data2_batchadjusted_vst_corrected: Z:/omics/mRNAseq_FIMM/raw_counts_Deseq_vst_corrected.csv
  mRNAseq_sample_metadata: Z:/omics/mRNAseq_FIMM/metadata.tsv
  mRNAseq_feature_metadata: Z:/omics/mRNAseq_FIMM/gene-annotation.tsv
  mRNAseq_feature_metadata_vst: Z:/omics/mRNAseq_FIMM/mrna_vst_mapping.csv
  
  EMseq_SE_celltypeadjusted_random: Z:/omics/EM-seq_UU/filtered_datasets/adjusted_filtered_rand.RDS
  EMseq_SE_celltypeadjusted_mostvariable: Z:/omics/EM-seq_UU/filtered_datasets/adjusted_filtered_sd.RDS
  EMseq_SE_random: Z:/omics/EM-seq_UU/filtered_datasets/meth_filt_rand.RDS
  EMseq_SE_mostvariable: Z:/omics/EM-seq_UU/filtered_datasets/meth_filt_sd.RDS
  EMseq_Mvalues_celltypeadjusted_mostvariable: Z:/omics/EM-seq_UU/filtered_datasets/adjusted_filtered_sd_mval.RDS
  EMseq_Mvalues_celltypeadjusted_random: Z:/omics/EM-seq_UU/filtered_datasets/adjusted_filtered_rand_M_val.RDS
  
  h5_outdir: h5MAE
---

This R notebook prepares an `MultiAssayExperiment` [@Ramos2017] object based on *CSV* files containing preprocessed omics measurements, corresponding omics feature metadata (annotations of molecular entities; database identifers), sample measurement metadata (e.g., measurement date), and sample phenotype information.


```{r parameters}
print(Sys.Date())
print(params)
```


## Phenotype data

Read phenotype information from CSV file.

```{r read_phenotype_data}
pheno_df <- read.csv(params$phenotype_data, stringsAsFactors = TRUE)
row.names(pheno_df) <- pheno_df$subject.id
cat("Column names of pheno_df:\n")
print(names(pheno_df))
cat("Dimensions of pheno_df:\n")
dim(pheno_df)
```

Adjust order of factor levels for BMI and age.

```{r define_phenotype_factor_levels}
levels(pheno_df$BMI.group)
pheno_df$BMI.group <- factor(
  pheno_df$BMI.group, levels = c("healthy", "overweight", "obese"), 
  ordered = TRUE)

levels(pheno_df$Age.group)
pheno_df$Age.group <- factor(
  pheno_df$Age.group, levels = levels(pheno_df$Age.group), 
  ordered = TRUE)

age_pred <- readRDS('Z:/omics/EM-seq_UU/analysis/Age_prediction/age.RDS')


BLUP_complete <- sapply(pheno_df$subject.id, function(x) {
  if (x %in% age_pred$id) {
    value <- age_pred[age_pred$id == x, 26]
  }
  else {
    value <- NA
  }
  value
})

BLUP_ACC_complete <- sapply(pheno_df$subject.id, function(x) {
  if (x %in% age_pred$id) {
    value <- age_pred[age_pred$id == x, 27]
  }
  else {
    value <- NA
  }
  value
})

Levin_complete <- sapply(pheno_df$subject.id, function(x) {
  if (x %in% age_pred$id) {
    value <- age_pred[age_pred$id == x, 10]
  }
  else {
    value <- NA
  }
  value
})

Levin_acc_complete <- sapply(pheno_df$subject.id, function(x) {
  if (x %in% age_pred$id) {
    value <- age_pred[age_pred$id == x, 11]
  }
  else {
    value <- NA
  }
  value
})
#rownames(age_pred) < age_pred$id
pheno_df$Age_pred_BLUP <-BLUP_complete
pheno_df$Age_acc_BLUP <- BLUP_ACC_complete
pheno_df$Age_pred_Levine <-Levin_complete
pheno_df$Age_acc_Levine <-Levin_acc_complete
```

## Create SummarizedExperiment per omics

The R/Bioconductor package `MultiAssayExperiment` is based on `SummarizedExperiment` [@Morgan2022]. `SummarizedExperiment` provides a container for one or several assays performed on the same set of samples. Since sample sizes vary per omics in the EATRIS-Plus multi-omics demonstrator cohort. One `SummarizedExperiment` is created per assay. In addition to the omics profiles, the object contains metadata about features (`rowData`) and samples (`colData`). Metadata can be added to the entire object. We use this to provide a short summary of the applied data processing methods (filtering, batch effect adjustment, missing value imputation, normalization).

The `SummarizedExperiment`s are collect in a `MultiAssayExperiment::ExperimentList`, which can be used in input for creating the `MultiAssayExperiment`.

```{r}
library(SummarizedExperiment)
library(MultiAssayExperiment)
library(HDF5Array)
```


### Metabolomics data and metadata

Prepare several `SummarizedExperiment` objects for the targeted metabolomics assays (amino acids, acylcarnitines, very long chain fatty acids). 

Names of experiments:

| Assay | Data without batch-effect adjustment or missing value imputation | Batch-adjusted data | Batch-adjusted and missing value imputed |
|-------|------------------------------------------------------------------|---------------------|------------------------------------------|
| Acylcarnitines | `Acylcarnitines \| with missing values` | `Acylcarnitines \| batch-adjusted \| with missing values` | `Acylcarnitines \| batch-adjusted \| imputed missing values` |
| Amino acids | `Amino acids \| with missing values` | `Amino acids \| batch-adjusted \| with missing values` | `Amino acids \| batch-adjusted \| imputed missing values` |
| Very long chain fatty acids | `Very long chain fatty acids` | `Very long chain fatty acids \| batch-adjusted` | Not applicable (data has no missing values) |


```{r prepare_metabolomics_experiments}
experiment_list <- ExperimentList(
  lapply(list(
    
    `Acylcarnitines | with missing values` = list(
      assay_name = "Acylcarnitines | with missing values",
      assay_data = params$acylcarnitines_data,
      assay_sample_metadata = params$acylcarnitines_sample_metadata,
      assay_feature_metadata = params$acylcarnitines_feature_metadata,
      batch_effect_correction = "No batch effect adjustment",
      missing_value_imputation = "Missing values were not imputed"),
    
    `Acylcarnitines | unscaled` = list(
      assay_name = "Acylcarnitines | unscaled",
      assay_data = params$acylcarnitines_data_unscaled,
      assay_sample_metadata = params$acylcarnitines_sample_metadata,
      assay_feature_metadata = params$acylcarnitines_feature_metadata,
      batch_effect_correction = "No batch effect adjustment",
      missing_value_imputation = "Missing values were not imputed"),
    
    `Amino acids | with missing values` = list(
      assay_name = "Amino acids | with missing values",
      assay_data = params$aminoacids_data,
      assay_sample_metadata = params$aminoacids_sample_metadata,
      assay_feature_metadata = params$aminoacids_feature_metadata,
      batch_effect_correction = "No batch effect adjustment",
      missing_value_imputation = "Missing values were not imputed"),
    
    `Amino acids | unscaled` = list(
      assay_name = "Amino acids | unscaled",
      assay_data = params$aminoacids_data_unscaled,
      assay_sample_metadata = params$aminoacids_sample_metadata,
      assay_feature_metadata = params$aminoacids_feature_metadata,
      batch_effect_correction = "No batch effect adjustment",
      missing_value_imputation = "Missing values were not imputed"),
    
    `Very long chain fatty acids` = list(
      assay_name = "Very long chain fatty acids",
      assay_data = params$fattyacids_data,
      assay_sample_metadata = params$fattyacids_sample_metadata,
      assay_feature_metadata = params$fattyacids_feature_metadata,
      batch_effect_correction = "No batch effect adjustment",
      missing_value_imputation = "Missing values were not imputed"),
    
    `Very long chain fatty acids | unscaled` = list(
      assay_name = "Very long chain fatty acids | unscaled",
      assay_data = params$fattyacids_data_unscaled,
      assay_sample_metadata = params$fattyacids_sample_metadata,
      assay_feature_metadata = params$fattyacids_feature_metadata,
      batch_effect_correction = "No batch effect adjustment",
      missing_value_imputation = "Missing values were not imputed"),
    
    # batch effect adjusted
    `Acylcarnitines | batch-adjusted | with missing values` = list(
      assay_name = "Acylcarnitines | batch-adjusted | with missing values",
      assay_data = params$acylcarnitines_data_batchadjusted,
      assay_sample_metadata = params$acylcarnitines_sample_metadata,
      assay_feature_metadata = params$acylcarnitines_feature_metadata,
      batch_effect_correction = "Batch effect adjustment was performed with ComBat using the parametric method. The variable 'Sex' was used as covariate",
      missing_value_imputation = "Missing values were not imputed"),
    
    `Amino acids | batch-adjusted | with missing values` = list(
      assay_name = "Amino acids | batch-adjusted | with missing values",
      assay_data = params$aminoacids_data_batchadjusted,
      assay_sample_metadata = params$aminoacids_sample_metadata,
      assay_feature_metadata = params$aminoacids_feature_metadata,
      batch_effect_correction = "Batch effect adjustment was performed with ComBat using the parametric method. The variable 'Sex' was used as covariate",
      missing_value_imputation = "Missing values were not imputed"),
    
    `Very long chain fatty acids | batch-adjusted` = list(
      assay_name = "Very long chain fatty acids | batch-adjusted",
      assay_data = params$fattyacids_data_batchadjusted,
      assay_sample_metadata = params$fattyacids_sample_metadata,
      assay_feature_metadata = params$fattyacids_feature_metadata,
      batch_effect_correction = "Batch effect adjustment was performed with ComBat using the parametric method. The variable 'Sex' was used as covariate",
      missing_value_imputation = "Missing values were not imputed"),
    
    `Acylcarnitines | batch-adjusted | imputed missing values` = list(
      assay_name = "Acylcarnitines | batch-adjusted | imputed missing values",
      assay_data = params$acylcarnitines_data_batchadjusted_imputed,
      assay_sample_metadata = params$acylcarnitines_sample_metadata,
      assay_feature_metadata = params$acylcarnitines_feature_metadata,
      batch_effect_correction = "Batch effect adjustment was performed with ComBat using the parametric method. The variable 'Sex' was used as covariate",
      missing_value_imputation = "Low abundance resampling"),
    
    `Amino acids | batch-adjusted | imputed missing values` = list(
      assay_name = "Amino acids | batch-adjusted | imputed missing values",
      assay_data = params$aminoacids_data_batchadjusted_imputed,
      assay_sample_metadata = params$aminoacids_sample_metadata,
      assay_feature_metadata = params$aminoacids_feature_metadata,
      batch_effect_correction = "Batch effect adjustment was performed with ComBat using the parametric method. The variable 'Sex' was used as covariate",
      missing_value_imputation = "Low abundance resampling")
    
  ), function(a) {
    # data
    my_assay = read.csv(a$assay_data, row.names = 1)
    my_assay_list <- list(my_assay)
    names(my_assay_list) <- c(a$assay_name)
    # sample metadata
    sample_metadata <- read.csv(
      a$assay_sample_metadata, stringsAsFactors = TRUE)
    rownames(sample_metadata) <- sample_metadata$subject.id
    sample_metadata <- sample_metadata[colnames(my_assay), ]
    # feature metadata
    feature_metadata <- read.delim(a$assay_feature_metadata)
    rownames(feature_metadata) <- feature_metadata$metabolite_identification
    selected_metadata_columns <- c(
      "database_identifier", "metabolite_identification", "chemical_formula", 
      "smiles", "inchi","database", "database_version", "uri", "HMDB.ID.s.", 
      "Chebi.ID.s")
    feature_names <- rownames(my_assay)
    feature_metadata <- feature_metadata[
      feature_names, 
      intersect(names(feature_metadata), selected_metadata_columns)]
    # SummarizedExperiment    
    se <- SummarizedExperiment(
      assays = my_assay_list,
      # rows: features
      rowData = feature_metadata, 
      # columns: samples
      colData = sample_metadata, 
      # metadata about data processing
      metadata = list(
        filtering = "Features that are missing in more than 30% of the sample or in all samples or at least one of the measured batches were removed",
        log_transformation = "Metabolite levels x [µM concentrations] were transformed using the natural logarithm: log(x+1)",
        batch_effect_correction = a$batch_effect_correction,
        missing_value_imputation = a$missing_value_imputation,
        standardization = "All features were centered to mean = 0 and scaled to var = 1"
      ),
      checkDimnames = TRUE)
  })
)
```


### Lipidomics data and metadata

Lipidomics profiles are not provided matrix-type format with sample IDs in columns following naming "CZC<4-digit-ID>". Sample renaming is required and performed below.

Names of experiments:

| Assay | Pre-processed data |
|-------|--------------------|
| Lipidomics, positive ion-mode | `Lipidomics, positive` |
| Lipidomics, negative ion-mode | `Lipidomics, negative` |


```{r prepare_lipidomics}
lipidomics_experiments_list <- ExperimentList(
  lapply(list(
    `Lipidomics, positive | transformed` = list(
      assay_name = "Lipidomics, positive | transformed",
      assay_data = params$lipidomics_pos_data,
      assay_sample_metadata = params$lipidomics_pos_sample_metadata,
      assay_feature_metadata = params$lipidomics_pos_feature_metadata,
      ion_mode = "postive"),
    `Lipidomics, negative | transformed` = list(
      assay_name = "Lipidomics, negative | transformed",
      assay_data = params$lipidomics_neg_data,
      assay_sample_metadata = params$lipidomics_neg_sample_metadata,
      assay_feature_metadata = params$lipidomics_neg_feature_metadata,
      ion_mode = "negative")
  ), function(a) {
    my_assay = read.csv(a$assay_data, row.names = 1)
    cohort_samples <- colnames(my_assay)[grep("CZC", colnames(my_assay))]
    my_assay <- my_assay[, cohort_samples]
    my_assay_list <- list(my_assay)
    names(my_assay_list) <- c(a$assay_name)
    # sample metadata
    sample_metadata <- read.delim(
      a$assay_sample_metadata, stringsAsFactors = TRUE,
      row.names = 1)
    sample_metadata$subject.id <- rownames(sample_metadata)
    sample_metadata <- sample_metadata[colnames(my_assay), ]
    # feature metadata
    feature_metadata <- read.delim(a$assay_feature_metadata, 
                                   check.names = FALSE)
    rownames(feature_metadata) <- feature_metadata$IMTM_Feature_ID
    feature_names <- rownames(my_assay)
    feature_metadata <- feature_metadata[
      feature_names, ]
    # SummarizedExperiment    
    se <- SummarizedExperiment(
      assays = my_assay_list,
      # rows: features
      rowData = feature_metadata, 
      # columns: samples
      colData = sample_metadata, 
      # metadata about data processing
      metadata = list(
        filtering = paste0("Data processed in Compound Discoverer 3.3 SP1, filtered based on spectral matches agains LipidBlast library with score > 50;	only ", a$ion_mode, " ionisation mode"),
        retention_time_alignment = "Retention time aligment",
        grouping = "Compound detection and adduct grouping",
        gap_filling = "Gap filling is performed applying 'Real Peak detection' with lower intensity setting",
        background_substraction = "Background substration was performed based on 'Blank samples' with ration Max. Sample/Blank = 20",
        normalization = "MAD (Median-Absolute-Deviation) Normalisation"
      ),
      checkDimnames = TRUE)
  })
)

experiment_list <- append(experiment_list, lipidomics_experiments_list)

stopifnot(!any(is.na(assays(lipidomics_experiments_list$`Lipidomics, positive`)[[1]])))
stopifnot(!any(is.na(assays(lipidomics_experiments_list$`Lipidomics, negative`)[[1]])))
# no missing values present
```


### Lipidomics SPLINE data and metadata


```{r prepare_lipidomics-spline}
lipidomics_experiments_list <- ExperimentList(
  lapply(list(
    `Lipidomics-SPLINE | positive` = list(
      assay_name = "Lipidomics-SPLINE | positive",
      assay_data = params$lipidomics_spline_pos_data,
      assay_sample_metadata = params$lipidomics_spline_sample_metadata,
      assay_feature_metadata = params$lipidomics_spline_feature_metadata,
      ion_mode = "postive")
  ), function(a) {
    my_assay = read.csv(a$assay_data, row.names = 1)
    my_assay <- data.frame(t(my_assay))
    my_assay_list <- list(my_assay)
    names(my_assay_list) <- c(a$assay_name)
    # sample metadata
    sample_metadata <- read.csv(
      a$assay_sample_metadata, stringsAsFactors = TRUE,
      row.names = 1)
    sample_metadata$subject.id <- sample_metadata$EATRIS_IDs
    rownames(sample_metadata) <- sample_metadata$subject.id
    sample_metadata$EATRIS_IDs <- NULL
    sample_metadata <- sample_metadata[colnames(my_assay),]
    # feature metadata
    feature_metadata <- read.csv(a$assay_feature_metadata, 
                                   check.names = FALSE, 
                                 row.names = 1)
    feature_names <- rownames(my_assay)
    feature_metadata <- feature_metadata[
      feature_names, ]
    # SummarizedExperiment    
    se <- SummarizedExperiment(
      assays = my_assay_list,
      # rows: features
      rowData = feature_metadata, 
      # columns: samples
      colData = sample_metadata, 
      # metadata about data processing
      checkDimnames = TRUE)
  })
)

experiment_list <- append(experiment_list, lipidomics_experiments_list)
```


### Proteomics data and metadata

Files and metadata were retrieved from `Z:/omics/Proteomics_IMTM/README.md`. 

Names of experiments:

| Assay | Batch-adjusted data | Batch-adjusted and missing value imputed |
|-------|---------------------|------------------------------------------|
| Proteomics | `Proteomics \| with missing values` |  `Proteomics \| imputed missing values` |


```{r prepare_proteomics}

proteomics_experiments <- lapply(list(
  `Proteomics | with missing values` = list(
    assay_name = "Proteomics | with missing values",
    assay_data = params$proteomics_data,
    assay_sample_metadata = params$proteomics_sample_metadata,
    assay_feature_metadata = params$proteomics_feature_metadata,
    missing_value_imputation = "Missing values were not imputed"),
  
  `Proteomics | imputed missing values` = list(
    assay_name = "Proteomics | imputed missing values",
    assay_data = params$proteomics_data_imputed,
    assay_sample_metadata = params$proteomics_sample_metadata,
    assay_feature_metadata = params$proteomics_feature_metadata,
    missing_value_imputation = "Random forest (missForest)")
  
), function(a) {
  # data
  my_assay = read.delim(a$assay_data)
  my_assay_list <- list(my_assay)
  names(my_assay_list) <- c(a$assay_name)
  
  # sample metadata
  sample_metadata <- read.delim(
    a$assay_sample_metadata, row.names = "ID")
  sample_metadata$subject.id <- factor(rownames(sample_metadata))
  sample_metadata <- sample_metadata[colnames(my_assay), ]
  # feature metadata
  feature_metadata <- read.delim(a$assay_feature_metadata)
  rownames(feature_metadata) <- feature_metadata$Accession
  feature_names <- rownames(my_assay)
  feature_metadata <- feature_metadata[feature_names, ]
  
  se <- SummarizedExperiment(
    assays = my_assay_list,
    # rows: features
    rowData = feature_metadata, 
    # columns: samples
    colData = sample_metadata, 
    # metadata about data processing
    metadata = list(
      filtering = "Proteins completely not quantified at least in one batch are filtered out. Proteins with low confidence are filtered out. Proteins with more than 30% of missing values are filtered out.",
      log_transformation = "Protein levels were log2-transformed",
      aggregation_of_technical_replicates = "Replicate measurements were aggregated by mean value",
      batch_effect_correction = "Batch effect adjustment was performed with ComBat using the variable 'Sex' as covariate",
      missing_value_imputation = a$missing_value_imputation,
      standardization = "No standardization"
    ),
    checkDimnames = TRUE)
})

experiment_list <- append(experiment_list, proteomics_experiments)
```



### Proteomics IPP data and metadata

Files and metadata were retrieved from `Z:/omics/Proteomics_IMTM/README.md`. 

Names of experiments:

```{r prepare_proteomics_ipp}

proteomics_experiments <- lapply(list(
  `Proteomics-IPP` = list(
    assay_name = "Proteomics-IPP",
    assay_data = params$proteomics_ipp_data,
    assay_sample_metadata = params$proteomics_ipp_sample_metadata,
    assay_feature_metadata = params$proteomics_ipp_feature_metadata)
  
), function(a) {
  # data
  my_assay = read.delim(a$assay_data)
  my_assay_list <- list(my_assay)
  names(my_assay_list) <- c(a$assay_name)
  
  # sample metadata
  sample_metadata <- read.delim(
    a$assay_sample_metadata, row.names = "CZCid")
  sample_metadata$subject.id <- factor(rownames(sample_metadata))
  sample_metadata <- sample_metadata[colnames(my_assay), ]
  # feature metadata
  feature_metadata <- read.delim(a$assay_feature_metadata)
  rownames(feature_metadata) <- feature_metadata$AptName
  feature_names <- rownames(my_assay)
  feature_metadata <- feature_metadata[feature_names, ]
  
  se <- SummarizedExperiment(
    assays = my_assay_list,
    # rows: features
    rowData = feature_metadata, 
    # columns: samples
    colData = sample_metadata, 
    # metadata about data processing
    checkDimnames = TRUE)
})

experiment_list <- append(experiment_list, proteomics_experiments)
```



### mRNA-seq data and metadata

**NOTE:** processing seems to be incomplete. Batch effect adjusted should be performed.

Sample IDs in columns do not follow the naming "CZC<4-digit-ID>". Sample renaming is required and performed below.

Note that for mRNA-seq, `rowRanges` are used instead of `rowData` to report metadata on transcripts (genomic ranges).

Names of experiments:

| Assay | Normalized CPM counts | Normalized CPM counts, batch-adjusted |
|-------|-----------------------|---------------------------------------|
| mRNA-seq | `mRNA-seq` | `mRNA-seq | batch-adjusted` |

```{r prepare_mRNAseq}

mRNAseq_experiments <- lapply(list(
  
  `mRNA-seq-2 | batch-adjusted | vst` = list(
    assay_name = "mRNAseq_data2_batchadjusted_VST",
    assay_data = params$mRNAseq_data2_batchadjusted_vst,
    assay_sample_metadata = params$mRNAseq_sample_metadata,
    assay_feature_metadata = params$mRNAseq_feature_metadata
  ),
  `mRNA-seq-2 | batch-adjusted | vst | corrected` = list(
    assay_name = "mRNAseq_data2_batchadjusted_vst_corrected",
    assay_data = params$mRNAseq_data2_batchadjusted_vst_corrected,
    assay_sample_metadata = params$mRNAseq_sample_metadata,
    assay_feature_metadata = params$mRNAseq_feature_metadata
  )

), function(a) {
  print(a)
  # data
  my_assay = read.csv(a$assay_data, row.names = 1)
  my_assay = data.frame(my_assay)
  measurement_names <- colnames(my_assay)
  
  # assure that subject IDs are now the same as in phenotype data
  stopifnot(ncol(my_assay) == length(intersect(colnames(my_assay), 
                                               rownames(pheno_df))))
  my_assay_list <- list(my_assay)

  # feature metadata
  feature_metadata <- read.csv(params$mRNAseq_feature_metadata_vst)
  rownames(feature_metadata) <- feature_metadata$Ensembl_ID
  # sample metadata
  sample_metadata <- read.delim(
    params$mRNAseq_sample_metadata, row.names = "Sample")
  rownames(sample_metadata) <- gsub("X", "CZC", rownames(sample_metadata))
  sample_metadata$subject.id <- factor(rownames(sample_metadata))
  sample_metadata <- sample_metadata[colnames(my_assay), ]
  
  se <- SummarizedExperiment(
    assays = my_assay_list,
    # rows: features
    rowData = feature_metadata,
    #rowRanges = grl, #RangedSummarizedExperiment
    # columns: samples
    colData = sample_metadata, 
    # metadata about data processing
    metadata = list("normalized CPM counts"),
    checkDimnames = TRUE)
  
}  
  
)

experiment_list <- append(experiment_list, mRNAseq_experiments)

```


### miRNA-seq data and metadata

**NOTE:** processing seems to be incomplete. Batch effect adjusted should be performed.

Sample IDs in columns do not follow the naming "CZC<4-digit-ID>". Sample renaming is required and performed below.

Names of experiments:

| Assay | Normalized CPM counts (mature miRNA) |  Normalized CPM counts (mature miRNA), batch-adjusted | Normalized CPM counts (hairpin miRNA) |
|-------|--------------------------------------|-------------------------------------------------------|---------------------------------------|
| miRNA-seq | `miRNA-seq, mature` | `miRNA-seq, mature | batch-adjusted` | `miRNA-seq, hairpin` |


```{r prepare_miRNAseq}
miRNAseq_experiments <- lapply(list(
  `miRNA-seq, mature` = list(
    assay_name = "miRNA-seq, mature",
    assay_data = params$miRNAseq_mature_data,
    assay_sample_metadata = params$miRNAseq_sample_metadata
    # assay_feature_metadata = params$ # no file with feature metadata
  ),
  `miRNA-seq, mature | batch-adjusted` = list(
    assay_name = "miRNA-seq, mature | batch-adjusted",
    assay_data = params$miRNAseq_mature_data_batchadjusted,
    assay_sample_metadata = params$miRNAseq_sample_metadata
    # assay_feature_metadata = params$ # no file with feature metadata
  ),
  `miRNA-seq, mature | batch-adjusted | corrected` = list(
    assay_name = "miRNA-seq, mature | batch-adjusted | corrected",
    assay_data = params$miRNAseq_mature_data_batchadjusted_celltypeadjusted,
    assay_sample_metadata = params$miRNAseq_sample_metadata
    # assay_feature_metadata = params$ # no file with feature metadata
  )
  # ,
  # `miRNA-seq, hairpin` = list(
  #   assay_name = "miRNA-seq, hairpin",
  #   assay_data = params$miRNAseq_hairpin_data,
  #   assay_sample_metadata = params$miRNAseq_sample_metadata
  #   # assay_feature_metadata = params$ # no file with feature metadata
  # )
), function(a) {
  # data
  if (grepl('corrected',  a$assay_name, fixed=TRUE)){
    my_assay = read.csv(a$assay_data, row.names = 1)
    my_assay = data.frame(my_assay)
    measurement_names <- colnames(my_assay)
  }
  else if (grepl('batch-adjusted',  a$assay_name, fixed=TRUE)){
    my_assay = t(read.csv(a$assay_data, row.names = 1))
    my_assay = data.frame(my_assay)
    measurement_names <- colnames(my_assay)
  }
  else {
    my_assay = read.delim(a$assay_data, row.names = 1)  
    measurement_names <- data.frame(
    measurement.id = sub("X", "", colnames(my_assay)))
    # columns are measurement names; replace by cohort subject ID
    # names(my_assay) <- sapply(names(my_assay), function(colname) {
    # subject.id <- paste0("CZC", substr(colname, 2, 5))
    # # TODO: does the rest of the file name (e.g., 3818_S104_L002_R1_001.mature) 
    # # contain any relevant sample/measurement metadata? 
    # rownames(measurement_names) <- colnames(my_assay)
    #})
    colnames(my_assay) <- lapply(colnames(my_assay), function(x){
      subjectid <- paste0("CZC", substr(x, 2, 5))
    })
  }
  
  
  # assure that subject IDs are now the same as in phenotype data
  stopifnot(ncol(my_assay) == length(intersect(colnames(my_assay), 
                                               rownames(pheno_df))))
  my_assay_list <- list(my_assay)
  names(my_assay_list) <- c(a$assay_name)
  # sample metadata
  sample_metadata <- read.csv(
    a$assay_sample_metadata, stringsAsFactors = TRUE)
  rownames(sample_metadata) <- sapply(sample_metadata$Sample, function(s) {
    paste0("CZC", s)
  })
  sample_metadata <- sample_metadata[colnames(my_assay), ]  
  sample_metadata$subject.id <- rownames(sample_metadata)
  sample_metadata <- cbind(sample_metadata, measurement_names)
  sample_metadata <- sample_metadata[colnames(my_assay), 
                                     !(names(sample_metadata) == "Sample")]
  # feature metadata
  # hsa-let-7f-5p_MIMAT0000067_Homo_sapiens_let-7f-5p
  # hsa-miR-26a-5p_MIMAT0000082_Homo_sapiens_miR-26a-5p
  feature_metadata <- do.call(
    rbind, 
    lapply(rownames(my_assay), function(feature.id) {
      feature_list <- strsplit(feature.id, "_")
      data.frame(
        feature.id = feature.id,
        miRBase.id = feature_list[[1]][1],
        miRBase.accession = feature_list[[1]][2])
    }))
  rownames(feature_metadata) <- feature_metadata$feature.id

  se <- SummarizedExperiment(
    assays = my_assay_list,
    # rows: features
    rowData = feature_metadata, 
    # columns: samples
    colData = sample_metadata, 
    # metadata about data processing
    metadata = list(
      preprocessing = "normalized count per million (CPM)"
    ),
    checkDimnames = TRUE)
})

experiment_list <- append(experiment_list, miRNAseq_experiments)
```


### miRNA qRT-PCR data and metadata

**NOTE:** information about processing (metadata) needs to be added.

| Assay | Processed data |
|-------|----------------|
| miRNA qRT-PCR profiling | `miRNA qRT-PCR` |

```{r}
my_assay_list = list(
  `miRNA qRT-PCR` = read.delim(params$mRNAqRTPCR_data, row.names = "miRNA"))

sample_metadata <- read.delim(params$mRNAqRTPCR_sample_metadata)
sample_metadata <- sample_metadata[ !(is.na(sample_metadata$RT)),]
rownames(sample_metadata) <- gsub("CZG", "CZC", sample_metadata$ID.Patient)
sample_metadata <- sample_metadata[colnames(my_assay_list[[1]]), ] 


se <- SummarizedExperiment(
    assays = my_assay_list,
    # # rows: features
    # rowData = feature_metadata, 
    # # columns: samples
    colData = sample_metadata, 
    # # metadata about data processing
    # metadata = list(),
    checkDimnames = TRUE)

experiment_list <- append(experiment_list, list(`miRNA qRT-PCR` = se))
```


### aCGH data and metadata

```{r}
my_assay <- read.csv(params$aCGH_data, row.names = 1)

my_assay_list = list(
  `aCGH` = read.csv(params$aCGH_data, row.names = 1))

se <- SummarizedExperiment(
    assays = my_assay_list,
    # # rows: features
    #rowData = feature_metadata, 
    # # columns: samples
    #colData = sample_metadata, 
    # # metadata about data processing
    # metadata = list(),
    checkDimnames = TRUE)

experiment_list <- append(experiment_list, list(`aCGH` = se))
```


### EM-seq data and metadata

Beta values are reported per CpG site. Multiple data set are provided based on selection of CpG sites and cell-type adjustment. See the README on EM-seq data processing for more information.

**NOTE:** the input files need to be updated with the latest version of files. Check the README on EM-seq data processing from Reka.

```{r prepare_EMseq}
require(methrix)

EMseq_data <- list(
  # "methrix" - assay -> methylation (beta) and coverage (cov)
  `EM-seq | 100,000 random CpG sites` = readRDS(params$EMseq_SE_random), 
  `EM-seq | 100,000 most variable CpG sites` = readRDS(params$EMseq_SE_mostvariable),
  # "matrix" "array" 
  `EM-seq | cell-type adjusted | 100,000 random CpG sites` = readRDS(params$EMseq_SE_celltypeadjusted_random),
  # "data.frame"
  `EM-seq | cell-type adjusted | 100,000 most variable CpG sites` = readRDS(params$EMseq_SE_celltypeadjusted_mostvariable),
  `EM-seq | M-values | cell-type adjusted | 100,000 most variable CpG sites` = readRDS(params$EMseq_Mvalues_celltypeadjusted_mostvariable),
  `EM-seq | M-values | cell-type adjusted | 100,000 random CpG sites` = readRDS(params$EMseq_Mvalues_celltypeadjusted_random)
  )

EMseq_experiments <- lapply(EMseq_data[1:2], function(x) {
  # "methrix" - assay -> methylation (beta) and coverage (cov)
  row_names <- apply(rowData(x), 1, function(row) {
    paste0(row[["chr"]], "_", row[["start"]]) 
  })
  assay_list <- lapply(assays(x), function(a) {
    rownames(a) <- row_names
    a 
  })
  feature_metadata <- rowData(x)
  rownames(feature_metadata) <- row_names
  SummarizedExperiment(
     assays = assay_list,
    # rows: features
    rowData = feature_metadata, 
    # columns: samples
    colData = colData(EMseq_data$`EM-seq | 100,000 random CpG sites`), 
    # metadata about data processing
    metadata = list(
      cell_type_adjustment = "no cell type adjustment"),
    checkDimnames = TRUE)
})

# # "matrix" "array" 
# `EM-seq | cell-type adjusted | 100,000 random CpG sites` = readRDS(params$EMseq_SE_celltypeadjusted_random), 
colnames(EMseq_data$`EM-seq | cell-type adjusted | 100,000 random CpG sites`) <- rownames(colData(EMseq_data$`EM-seq | 100,000 random CpG sites`))
assay_list <- list(
  `EM-seq | cell-type adjusted | 100,000 random CpG sites` = data.frame(
    EMseq_data$`EM-seq | cell-type adjusted | 100,000 random CpG sites`)
)
feature_metadata <- data.frame(
  chr = sapply(
    rownames(EMseq_data$`EM-seq | cell-type adjusted | 100,000 random CpG sites`), 
    function(x) {strsplit(x, "_")[[1]][1]}),
  start = sapply(
    rownames(EMseq_data$`EM-seq | cell-type adjusted | 100,000 random CpG sites`), 
    function(x) {strsplit(x, "_")[[1]][2]}),
  strand = "*"
)
EMseq_experiments$`EM-seq | cell-type adjusted | 100,000 random CpG sites` <- SummarizedExperiment(
  assays = assay_list,
  # rows: features
  rowData = feature_metadata, 
  # metadata about data processing
  metadata = list(
    cell_type_adjustment = "beta-values were adjusted for cell-type composition"),
  checkDimnames = TRUE)

# # "data.frame"
# `EM-seq | cell-type adjusted | 100,000 most variable CpG sites` = readRDS(params$EMseq_SE_celltypeadjusted_mostvariable)
assay_list <- list(
  `EM-seq | cell-type adjusted | 100,000 most variable CpG sites` =
    EMseq_data$`EM-seq | cell-type adjusted | 100,000 most variable CpG sites`
)
feature_metadata <- data.frame(
  chr = sapply(
    rownames(EMseq_data$`EM-seq | cell-type adjusted | 100,000 most variable CpG sites`),
    function(x) {strsplit(x, "_")[[1]][1]}),
  start = sapply(
    rownames(EMseq_data$`EM-seq | cell-type adjusted | 100,000 most variable CpG sites`),
    function(x) {strsplit(x, "_")[[1]][2]}),
  strand = "*"
)
EMseq_experiments$`EM-seq | cell-type adjusted | 100,000 most variable CpG sites` <- SummarizedExperiment(
  assays = assay_list,
  # rows: features
  rowData = feature_metadata,
  # metadata about data processing
  metadata = list(
    cell_type_adjustment = "beta-values were adjusted for cell-type composition"),
  checkDimnames = TRUE)



# # "data.frame"
# `EM-seq | M-values | cell-type adjusted | 100,000 most variable CpG sites` = readRDS(params$EMseq_Mvalues_celltypeadjusted_mostvariable)
# assay_list <- list(
#   `EM-seq | M-values | cell-type adjusted | 100,000 most variable CpG sites` =
#     EMseq_data$`EM-seq | M-values | cell-type adjusted | 100,000 most variable CpG sites`
# )
# feature_metadata <- data.frame(
#   chr = sapply(
#     rownames(EMseq_data$`EM-seq | M-values | cell-type adjusted | 100,000 most variable CpG sites`),
#     function(x) {strsplit(x, "_")[[1]][1]}),
#   start = sapply(
#     rownames(EMseq_data$`EM-seq | M-values | cell-type adjusted | 100,000 most variable CpG sites`),
#     function(x) {strsplit(x, "_")[[1]][2]}),
#   strand = "*"
# )
# EMseq_experiments$`EM-seq | M-values | cell-type adjusted | 100,000 most variable CpG sites` <- SummarizedExperiment(
#   assays = assay_list,
#   # rows: features
#   rowData = feature_metadata,
#   # metadata about data processing
#   metadata = list(
#     cell_type_adjustment = "beta-values were adjusted for cell-type composition"),
#   checkDimnames = TRUE)



# # "data.frame"
# `EM-seq | M-values | cell-type adjusted | 100,000 most variable CpG sites` = readRDS(params$EMseq_Mvalues_celltypeadjusted_mostvariable)
# assay_list <- list(
#   `EM-seq | M-values | cell-type adjusted | 100,000 random CpG sites` =
#     EMseq_data$`EM-seq | M-values | cell-type adjusted | 100,000 random CpG sites`
# )
# feature_metadata <- data.frame(
#   chr = sapply(
#     rownames(EMseq_data$`EM-seq | M-values | cell-type adjusted | 100,000 random CpG sites`),
#     function(x) {strsplit(x, "_")[[1]][1]}),
#   start = sapply(
#     rownames(EMseq_data$`EM-seq | M-values | cell-type adjusted | 100,000 random CpG sites`),
#     function(x) {strsplit(x, "_")[[1]][2]}),
#   strand = "*"
# )
# EMseq_experiments$`EM-seq | M-values | cell-type adjusted | 100,000 random CpG sites` <- SummarizedExperiment(
#   assays = assay_list,
#   # rows: features
#   rowData = feature_metadata,
#   # metadata about data processing
#   metadata = list(
#     cell_type_adjustment = "beta-values were adjusted for cell-type composition"),
#   checkDimnames = TRUE)


experiment_list <- append(experiment_list, EMseq_experiments)
```


## Create MultiAssayExperiment

The `SummarizedExperiment`s are now added to a `MultiAssayExperiment`. In addition to the `ExperimentList`, phenotype information on individuals is added (colData). General metadata about the cohort can be added here.

```{r create_multiassayexperiment}
mae <- MultiAssayExperiment(
  experiments = experiment_list,
  colData = pheno_df,
  metadata = list(
    description = "Multi-omics data of EATRIS-Plus healthy Czech demonstrator cohort"
  )
)

mae
```


## Remove mixup samples
```{r}
sample_names <- rownames(colData(mae))
sample_names_NoMixup <- sample_names[!sample_names %in% c("CZC2898", "CZC2935")]

mae <- subsetByColData(mae, y = sample_names_NoMixup)
mae
```  

Save MultiAssayExperiment to HDF5 file.

```{r save_multiassayexperiment}
saveHDF5MultiAssayExperiment(mae, dir = params$h5_outdir, replace = TRUE)
```


## Session information

```{r session_info}
sessionInfo()
```

## References
