---
title: 'Perform mRNA data preprocessing (batch correction on raw counts)'
author:
- Casper de Visser^[Radboud university medical center, Casper.deVisser@radboudumc.nl]
output:
  html_document:
    toc: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### File paths
```{r}
mRNA_df_path <- 'Z:/omics/mRNAseq_FIMM/gene-counts.txt'
mRNA_metadata_path <- 'Z:/omics/mRNAseq_FIMM/metadata.tsv'
phenotype_file <- "Z:/omics/Phenotype/phenotypes_from_phenopackets_2022-12-01.csv"
anno_file <- 'Z:/omics/mRNAseq_FIMM/gene-annotation.tsv' 
```

### Read annotationfile
```{r}
anno <- read.table(anno_file, sep='\t', header=T)
```

### Read mRNA data into dataframe

```{r}
mRNA_df <- read.csv(mRNA_df_path, sep = "\t", row.names=1)
mRNA_df <- mRNA_df[, 1:92]
colnames(mRNA_df) <- gsub("X", "CZC", colnames(mRNA_df))

```



### Read metadata

```{r}
mRNA_metadata_df <- read.csv(mRNA_metadata_path, row.names='Sample', sep='\t')
rownames(mRNA_metadata_df) <- gsub("X", "CZC", rownames(mRNA_metadata_df))
mRNA_metadata_df$subject.id <- factor(rownames(mRNA_metadata_df))
mRNA_metadata_df <- mRNA_metadata_df[colnames(mRNA_df), ]

```

### Phenotype data

Read phenotype information from Phenopackets json file.

```{r}
pheno_df <- read.csv(phenotype_file)
row.names(pheno_df) <- pheno_df$subject.id
cat("Column names of pheno_df")
print(names(pheno_df))
cat("Dimensions of pheno_df")
dim(pheno_df)
```

Convert selected columns to factor

```{r}
pheno_factor_columns <- c("subject.id", "Sex", "Smoking.Status", 
                          "ABO.Blood.Group", "Rh.Blood.Group", 
                          "PhenotypicFeature", "Age.group", "BMI.group")
pheno_df[, pheno_factor_columns] <- lapply(
  pheno_df[, pheno_factor_columns], factor)
pheno_df$BMI.group <- factor(
  pheno_df$BMI.group, levels = c("healthy", "overweight", "obese"), 
  ordered = TRUE)
```

Remove samples that are not in mRNA data

```{r}
pheno_df <- pheno_df[colnames(mRNA_df), ]
```


## Batch effect correction with Combat
```{r}
library(sva)
citation("sva")
```

## Batch effect correction 

```{r}
cat("non-parametric / sex as covariate\n")
adjusted <- ComBat_seq(as.matrix(mRNA_df), batch= mRNA_metadata_df$Group, group=pheno_df$Sex)
```


## EdgeR

```{r}
count_dge <- edgeR::DGEList(counts = adjusted, genes = anno)
count_dge <- edgeR::calcNormFactors(count_dge)
count_dge$samples$group <- pheno_df$Sex
count_dge$samples
```

## Filter low expressed transripts
``` {r}
keep <- edgeR::filterByExpr(count_dge, group=count_dge$samples$group)
count_dge_filtered <- count_dge[keep, , keep.lib.sizes=FALSE]
```


```{r}
norm_counts <- edgeR::cpm(count_dge_filtered, normalized.lib.sizes=TRUE, log=TRUE)
dim(norm_counts)
head(norm_counts)
``` 


```{r}
#estimated_cell_proportions <- read.csv('Z:/omics/mRNAseq_FIMM/estimated_cell_proportions.csv', row.names = 1)
estimated_cell_proportions <- readRDS('Z:/omics/EM-seq_UU/analysis/scripts/cell_type_composition.RDS')

# Remove erythrocyte column
estimated_cell_proportions <- data.frame(estimated_cell_proportions[, 2:15])

# Subset samples
estimated_cell_proportions <- estimated_cell_proportions[colnames(norm_counts),]
```


## Fill missing values with median of each column
```{r}
estimated_cell_proportions_imputed <- missMethods::impute_median(estimated_cell_proportions)

# Add names of samples
rownames(estimated_cell_proportions_imputed) <- colnames(norm_counts)
estimated_cell_proportions <- estimated_cell_proportions_imputed
```

## Correct microRNA-seq data for the estimated cell type proportions

```{r}
# 
mrna_seq_correct <- resid(lm(t(norm_counts) ~ estimated_cell_proportions$Blood.T.CD3 +
                                        estimated_cell_proportions$Blood.T.CD4 +
                                        estimated_cell_proportions$Blood.T.CD8 +
                                        estimated_cell_proportions$Blood.T.CenMem.CD4 +
                                        estimated_cell_proportions$Blood.T.Eff.CD8 +
                                        estimated_cell_proportions$Blood.T.EffMem.CD8 +
                                        estimated_cell_proportions$Blood.T.Naive.CD4 +
                                        estimated_cell_proportions$Blood.T.Naive.CD8 +
                                        estimated_cell_proportions$Blood.NK +
                                        estimated_cell_proportions$Blood.Monocytes +
                                        estimated_cell_proportions$Blood.Granulocytes +
                                        estimated_cell_proportions$Blood.B +
                                        estimated_cell_proportions$Blood.B.Mem))

mrna_seq_correct <- t(data.frame(mrna_seq_correct))
```

## Plot differences
```{r}
# boxplot(norm_counts)
boxplot(mrna_seq_correct)
# boxplot(scale(mrna_seq_correct))
# boxplot(-log(mrna_seq_correct))
```


## Write to CSV

```{r}
write.csv(t(norm_counts), 'Z:/omics/mRNAseq_FIMM/raw_counts_CombatSeq_cpm.csv')
write.csv(mrna_seq_correct, 'Z:/omics/mRNAseq_FIMM/raw_counts_CombatSeq_cpm_corrected.csv')
```




# Deseq2
```{r}
library(DESeq2)


dds <- DESeqDataSetFromMatrix(countData = adjusted, colData = pheno_df, design = ~ Sex)

dds <- DESeq(dds)

smallest_groupsize <- 5
keep <- rowSums(counts(dds) >= 15) >= smallest_groupsize
dds <- dds[keep,]

vst <- vst(dds, blind = FALSE)

#data_rlog <- rlog(dds, blind=FALSE)

dim(assay(vst))
```


## Save to .csv
```{r}
vst_data <- assay(vst)
write.csv(vst_data, 'Z:/omics/mRNAseq_FIMM/raw_counts_Deseq_vst.csv')

#data_rlog <- assay(data_rlog)
#write.csv(data_rlog, 'Z:/omics/mRNAseq_FIMM/raw_counts_Deseq_rlog.csv')
```



### Cell type correct data preprocessed with DeSeq2
```{r}
corrected_vst_data <- resid(lm(t(vst_data) ~ estimated_cell_proportions$Blood.T.CD3 +
                                        estimated_cell_proportions$Blood.T.CD4 +
                                        estimated_cell_proportions$Blood.T.CD8 +
                                        estimated_cell_proportions$Blood.T.CenMem.CD4 +
                                        estimated_cell_proportions$Blood.T.Eff.CD8 +
                                        estimated_cell_proportions$Blood.T.EffMem.CD8 +
                                        estimated_cell_proportions$Blood.T.Naive.CD4 +
                                        estimated_cell_proportions$Blood.T.Naive.CD8 +
                                        estimated_cell_proportions$Blood.NK +
                                        estimated_cell_proportions$Blood.Monocytes +
                                        estimated_cell_proportions$Blood.Granulocytes +
                                        estimated_cell_proportions$Blood.B +
                                        estimated_cell_proportions$Blood.B.Mem))

# residuals <- resid(fit)
# corrected_vst_data <- t(vst_data) - residuals


boxplot(t(corrected_vst_data))
```


### Cell type correct data preprocessed with DeSeq2 ( rlog() )
```{r}
corrected_rlog_data <- resid(lm(t(data_rlog) ~ estimated_cell_proportions$Blood.T.CD3 +
                                        estimated_cell_proportions$Blood.T.CD4 +
                                        estimated_cell_proportions$Blood.T.CD8 +
                                        estimated_cell_proportions$Blood.T.CenMem.CD4 +
                                        estimated_cell_proportions$Blood.T.Eff.CD8 +
                                        estimated_cell_proportions$Blood.T.EffMem.CD8 +
                                        estimated_cell_proportions$Blood.T.Naive.CD4 +
                                        estimated_cell_proportions$Blood.T.Naive.CD8 +
                                        estimated_cell_proportions$Blood.NK +
                                        estimated_cell_proportions$Blood.Monocytes +
                                        estimated_cell_proportions$Blood.Granulocytes +
                                        estimated_cell_proportions$Blood.B +
                                        estimated_cell_proportions$Blood.B.Mem))

# residuals <- resid(fit)
# corrected_vst_data <- t(vst_data) - residuals


boxplot(t(corrected_rlog_data))
```




## Write DeSeq2 Corrected to csv
```{r}
write.csv(t(corrected_vst_data), 'Z:/omics/mRNAseq_FIMM/raw_counts_DeSeq_vst_corrected.csv')

#write.csv(corrected_rlog_data, 'Z:/omics/mRNAseq_FIMM/raw_counts_DeSeq_rlog_corrected.csv')
```



