---
title: 'Perform miRNA data preprocessing (batch correction)'
author:
- Casper de Visser^[Radboud university medical center, Casper.deVisser@radboudumc.nl]
output:
  html_document:
    toc: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### File paths
```{r}
microRNA_df_path <- 'Z:/omics/microRNAseq_FIMM/miRBase_mature/mature_logtpm.csv'
microRNA_metadata_path <- 'Z:/omics/microRNAseq_FIMM/metadata.csv'
phenotype_file <- "Z:/omics/Phenotype/phenotypes_from_phenopackets_2022-12-01.csv"
```


### Read microRNA data 

```{r}
microRNA_df = read.csv(microRNA_df_path, row.names = 1)
measurement_names <- data.frame(
  measurement.id = sub("X", "", colnames(microRNA_df)))
  # columns are measurement names; replace by cohort subject ID
rownames(microRNA_df) <- lapply(rownames(microRNA_df), function(colname) {
   subject.id <- paste0("CZC", substr(colname, 1, 4))
 })
microRNA_df <- data.frame(t(microRNA_df))
```

## Read metadata

```{r}
microRNA_metadata_df <- read.csv(microRNA_metadata_path, stringsAsFactors = TRUE)
microRNA_metadata_df$Batch <- as.factor(microRNA_metadata_df$Batch)
rownames(microRNA_metadata_df) <- sapply(microRNA_metadata_df$Sample, function(s) {
    paste0("CZC", s) 
  })
```

### Sample in batch 6 is not included in Combat (as this it is not possible to have batch with one sample)
```{r}
sample_not_included <- rownames(microRNA_metadata_df[microRNA_metadata_df$Batch==6,])
microRNA_combat <- microRNA_df
microRNA_combat[sample_not_included] <- NULL

# Subset of metadata without batch 6
microRNA_metadata_combat <- microRNA_metadata_df[colnames(microRNA_combat), ]
```

### Phenotype data

Read phenotype information from Phenopackets json file.

```{r}
pheno_df <- read.csv(phenotype_file)
row.names(pheno_df) <- pheno_df$subject.id
cat("Column names of pheno_df")
print(names(pheno_df))
cat("Dimensions of pheno_df")
dim(pheno_df)
```

Convert selected columns to factor

```{r}
pheno_factor_columns <- c("subject.id", "Sex", "Smoking.Status", 
                          "ABO.Blood.Group", "Rh.Blood.Group", 
                          "PhenotypicFeature", "Age.group", "BMI.group")
pheno_df[, pheno_factor_columns] <- lapply(
  pheno_df[, pheno_factor_columns], factor)
pheno_df$BMI.group <- factor(
  pheno_df$BMI.group, levels = c("healthy", "overweight", "obese"), 
  ordered = TRUE)
```

Remove samples that are not in mRNA data

```{r}
pheno_df <- pheno_df[colnames(microRNA_df), ]
pheno_df_combat <- pheno_df[colnames(microRNA_combat), ] #Without batch 6 
```


## Batch effect correction with Combat
```{r}
library(sva)
citation("sva")
```


```{r}
run_combat <- function(input_df, metadata_df, parametric = TRUE, 
                       include_covariate = FALSE) {
  unadjusted <- input_df
  print(dim(unadjusted))
  sample_names <- colnames(input_df)
  tech_var <- metadata_df[sample_names, ]
  batch <- as.factor(tech_var[sample_names, ][["Batch"]])
  bio_var <- pheno_df[sample_names, c("Sex", "Age")]
  
  if (include_covariate) {
    mod <- model.matrix(~Sex, data = bio_var)
  } else {
    mod <- NULL
  }
  
  # how many male/female samples per batch?
  print(table(data.frame( 
    Measurement.date = tech_var$Batch, Sex = bio_var$Sex )))
  
  # without sex as covariate
  combat_adjusted <- ComBat( # _nosex_par
    dat = unadjusted,
    batch = batch,
    mod = NULL, # without sex as covariate
    par.prior = parametric, # parametric adjustment
    prior.plots = TRUE,
    mean.only = FALSE
  )
  
  as.data.frame(t(combat_adjusted))
}
```

## Including Plots

```{r}
cat("non-parametric / sex as covariate\n")
microRNA_combatCovSex <- run_combat(microRNA_combat, droplevels(microRNA_metadata_combat),  parametric = TRUE, include_covariate = TRUE)
```


## Add sample from batch 6 to batch-adjusted data

```{r}
microRNA_combatCovSex <- data.frame(t(microRNA_combatCovSex))
microRNA_combatCovSex[sample_not_included] <- microRNA_df[sample_not_included]
microRNA_combatCovSex <- data.frame(t(microRNA_combatCovSex))
```

#### Visualize data before and after batch effect correction

Principal Component Analysis score plot

```{r}
library(ggokabeito) # Okabe-Ito color palette
library(ggplot2)
library(stringr)
```

```{r}
num_PCs <- 5

data_list <- list(
  `Unadjusted` = t(microRNA_df),
  `ComBat-adjusted, parametric, sex as covariate` = microRNA_combatCovSex)
  
# run PCA and combine data for plots
plot_data <- do.call(rbind, lapply(names(data_list), function(adjustmethod) { df <- data_list[[adjustmethod]]
  # exclude features with missing values from PCA
  df <- df[, apply(df, 2, function(col) {!any(is.na(col))})]
  # perform PCA
  pca <- prcomp(df, center = TRUE, scale = TRUE)
  # PCA scores 
  pca_scores <- as.data.frame(pca$x[, 1:num_PCs])
  # add additional variables needed for plotting - batch, sex
  pca_scores$method <- as.factor(adjustmethod)
    pca_scores$Sex <- sapply(
    rownames(pca_scores), function(row_name) {
      pheno_df[row_name, "Sex"]})
  pca_scores$Batch <- sapply(
    rownames(pca_scores), function(row_name) {
      microRNA_metadata_df[row_name, "Batch"]})
  pca_scores
  }))
  
#plot PCA scores, solor by measurement date
ggp_batch <- ggplot(
  plot_data, 
  aes(x = PC1, y = PC2, 
      col = Batch, shape = Batch)) +
  geom_point() +
  stat_ellipse() +
  facet_wrap(.~method) +
  scale_color_okabe_ito() +
  ggtitle("PCA score plot, before/after batch-effect correction")

# plot PCA scores, solor by measurement sex
ggp_sex <- ggplot(
  plot_data, 
  aes(x = PC1, y = PC2, col = Sex, shape = Sex)) +
  geom_point() +
  stat_ellipse() +
  facet_wrap(.~method) +
  scale_color_okabe_ito() +
  ggtitle("PCA score plot, before/after batch-effect correction")

list(ggp_batch, ggp_sex)
```

## Same plot with different package shows all samples

### Original data

```{r}

library(ggfortify)

# perform PCA
df <- data_list[[1]]
df <- data.frame(df)
pca <- prcomp(df, center = TRUE, scale = TRUE)
# PCA scores 
df$Batch <- microRNA_metadata_df$Batch
autoplot(pca, data = df, colour = 'Batch', frame =TRUE, frame.colour='Batch')
``` 

### Batch-adjusted data

```{r}

library(ggfortify)

# perform PCA
df <- data_list[[2]]
df <- data.frame(df)
pca <- prcomp(df, center = TRUE, scale = TRUE)
# PCA scores 
df$Batch <- microRNA_metadata_df$Batch
autoplot(pca, data = df, colour = 'Batch', frame =TRUE, frame.colour='Batch')
``` 


PCA suggests the sample in batch 6 is an outlier. Perhaps because it has not been treated with Combat.
Therefore it is removed from the dataframe:

```{r}
microRNA_combatCovSex <- data.frame(t(microRNA_combatCovSex))
microRNA_combatCovSex[sample_not_included] <- NULL
microRNA_combatCovSex <- data.frame(t(microRNA_combatCovSex))
```


## Save batch corrected data to csv
```{r}
write.csv(microRNA_combatCovSex, 'Z:/omics/microRNAseq_FIMM/miRBase_mature/mature_Log_Combat_CovSex.csv')
```
